
# alte Analysen mit lmer

Im folgenden verwenden wir drei Kontraste: (0) Intercept = Gesamtdurchschnitt aller Gruppen, (1) Exp-Ø (MINT+Sprache)/2 vs. Control, (2) Sprache vs. MINT. Mit der Zeit-Interaktion testen wir, ob sich diese Unterschiede über die Zeit unterschiedlich entwickeln (Difference-in-Differences: Exp-Ø vs. Control sowie Sprache vs. MINT).
``` {r lmer MINT vs Sprache vs Kontrollgruppe}
#| eval: false
data <- data %>% droplevels() # drop unused factor levels

## define contrast Control vs Exp1+Exp2 - vermutlich bessere version, da die Estimates "stimmen"
mat <- cbind(
  c(1 / 3, 1 / 3, 1 / 3), # Durchschnitt aller 3 Gruppen
  c(-1, .5, .5), # Vergleich Control vs Mittel der beiden Interventions-Gruppen
  c(0, -1, 1)
) # Vergleich Sprache vs MINT
contrasts.CvsExp <- solve(t(mat))[, -1]
colnames(contrasts.CvsExp) <- c("-C+MINT_Sprache", "-MINT+Sprache")
contrasts(data$intervention) <- contrasts.CvsExp

# or use this coding system?
# my.helmert <- matrix(c(-2 / 3, 1 / 3, 1 / 3, 0, -1 / 3, 1 / 3), ncol = 2)
# colnames(my.helmert) <- c("-C+MINT_Sprache", "-MINT+Sprache")
# contrasts(data$cond2) <- my.helmert

## define variables
random_terms_lmer <- c("(1 | code)")
groups <- c("intervention")
outcomes <- c(
  "AILIT",
  "AILITWolke",
  "AILITShort",
  "TK",
  "TPK",
  "TCK",
  "TPACKce",
  "TPACKgenai",
  "Ethics",
  "PU",
  "PU1",
  "PU2",
  "PU3",
  "PU4"
)
time <- "time"
controls <- c("gender", "semester", "location")

# groups set as NA
lmers_prepost_all <-
  lapply(
    outcomes,
    fit_lmer,
    groups,
    time,
    controls,
    random_terms_lmer,
    NULL,
    data = data
  )

modelsummary_models(lmers_prepost_all)

```

```{r scale reliability with cronbach alpha}	
#| eval: false
## look at MSD and cronbach alphas for outcomes for each intervention group
# filter out pre data and relevant lists
data_pre <- data %>% filter(time == "pre")
all_scales <- c(
  list_items_i_tpack,
  list_items_pu
)
# 1) Cronbach's alpha for each scale
alphas <- imap(
  all_scales,
  ~ psych::alpha(dplyr::select(data_pre, dplyr::all_of(.x)), warnings = FALSE)
)

# 2) Extract only std.alpha
alpha_df <- tibble(
  scale = names(alphas),
  std_alpha = map_dbl(alphas, ~ .x$total$std.alpha)
)

# 3) Descriptives (composite = row means; switch to rowSums if you want sums)
desc_df <- imap_dfr(
  all_scales,
  ~ {
    comp <- rowMeans(dplyr::select(data_pre, dplyr::all_of(.x)), na.rm = TRUE)
    tibble(
      scale = .y,
      mean = mean(comp, na.rm = TRUE),
      sd = sd(comp, na.rm = TRUE),
      min = min(comp, na.rm = TRUE),
      max = max(comp, na.rm = TRUE)
    )
  }
)

# 4) Final summary: descriptives + std.alpha
result_df <- desc_df %>%
  left_join(alpha_df, by = "scale") %>%
  arrange(scale)

result_df

```

### Exploring Ref scale in posttest

```{r pca-ref}	
#| eval: false
data_post <- data %>% filter(time == "post" & Ref != "NaN")
## PCA
df_ref_items <- data_post %>%
  select(all_of(list_items_ref$Ref)) %>%
  na.omit()
fa.parallel(df_ref_items, fa = "pc", n.iter = 100, show.legend = FALSE) # tells us 3-4 components to extract
ncomp <- 3
pca_ref <- principal(
  df_ref_items,
  nfactors = ncomp,
  rotate = "none",
  scores = TRUE
)
print(pca_ref, digits = 3) # tells us that the 4 PCs explain 65% of variance (1st 31%, 2nd 14%, 3rd 12%, 4th 6.9%)

# rotate to get simpler structure
pca_ref_rot <- principal(
  df_ref_items,
  nfactors = ncomp,
  rotate = "varimax",
  scores = TRUE
)
print(pca_ref_rot, digits = 3, sort = TRUE)

# extract the loading matrix and look at each item
L <- as.data.frame(unclass(pca_ref_rot$loadings))

# find each item's dominant component (highest absolute loading)
L$primary <- colnames(L)[apply(abs(L), 1, which.max)]

# function to compute percentage contributions within each component
contrib_list <- lapply(unique(L$primary), function(comp) {
  sub <- L[L$primary == comp, ]
  load_abs2 <- (abs(sub[[comp]])^2)
  pct <- load_abs2 / sum(load_abs2) * 100
  data.frame(
    item = rownames(sub),
    component = comp,
    loading = round(sub[[comp]], 3),
    contrib_pct = round(pct, 2)
  )
})
contrib_tbl <- bind_rows(contrib_list) %>% arrange(component, desc(contrib_pct))
contrib_tbl
```

We see 3 strong components in the Ref scale. Further investigate the items with EFA and Cronbach's alpha.

```	{r ref-efa}
#| eval: false
ncomp <- 3
# EFA
df_ref_items <- data_post %>%
  select(all_of(list_items_ref$Ref)) %>%
  na.omit()
ref_scale <- list_items_ref %>% unlist(use.names = F)
# ref_scale <- c(
#   "RE01_01",
#   "RE01_03",
#   "RE01_04",
#   "RE01_05",
#   "RE01_06",
#   "RE01_07",
#   "RE01_11",
#   "RE01_13",
#   "RE01_16"
# )
# efa_varimax <- data_post %>%
#   select(any_of(ref_scale)) %>%
#   psych::fa(
#     nfactors = ncomp,
#     rotate = "varimax",
#     fm = "mlr",
#     ordered = F,
#     missing = T
#   )
# print(efa_varimax, digits = 2, cut = .3, sort = F)

efa_promax <- df_ref_items %>%
  select(any_of(ref_scale)) %>%
  psych::fa(nfactors = ncomp, rotate = "varimax", fm = "ml", missing = T)

# print(efa_promax, digits = 2, cut = .3, sort = F)
fa.diagram(efa_promax, digits = 2, cut = .30, simple = FALSE)

## Computing EFA using the lavaan package
tpack_efa <- lavaan::efa(
  df_ref_items %>%
    select(any_of(ref_scale)),
  nfactors = ncomp,
  rotation = "varimax",
  ordered = F,
  std.lv = TRUE,
  missing = "fiml"
)

# tpack_efa <- lavaan::efa(
#   df_ref_items %>%
#     select(any_of(ref_scale)),
#   nfactors = ncomp,
#   rotation = "none",
#   ordered = names(df_ref_items),
#   std.lv = TRUE,
#   estimator = "WLSMV"
# )

summary(tpack_efa, nd = 3L, cutoff = 0.3, dot.cutoff = 0.05)
# fitMeasures(tpack_efa, fit.measures = "all")
```	

items that could be removed: 
for scale 1: 02/09-11 have only weak or unclean loads; 16 is low loaded
for scale 3: 08 too weak and 15 too weak and cross-loading

```	{r ref-cronbach}
#| eval: false
## Alphas
scales <- list(
  core3 = c("RE01_01", "RE01_03", "RE01_13"),
  core4 = c("RE01_01", "RE01_03", "RE01_13", "RE01_12"),
  applied = c("RE01_04", "RE01_05", "RE01_14"),
  tech = c("RE01_06", "RE01_07"),
  total = ref_scale
)

alpha_summary <- imap_dfr(
  scales,
  ~ {
    dat <- df_ref_items %>%
      select(all_of(.x)) %>%
      mutate(across(everything(), ~ as.numeric(as.character(.))))
    a <- psych::alpha(dat, use = "pairwise", check.keys = TRUE)
    tibble(
      scale = .y,
      raw_alpha = round(a$total$raw_alpha, 3),
      std_alpha = round(a$total$std.alpha, 3),
      avg_r = round(a$total$average_r, 3),
      n_items = a$total$nvar
    )
  }
)

alpha_summary


```	


## Descriptive Plots

### Reshape data to long format

```{r reshape-long}
#| eval: false
data_long <- data %>%
  pivot_longer(
    cols = c(all_of(outcomes)),
    names_to = "outcome",
    values_to = "score"
  )

data_long <- data_long %>%
  mutate(
    outcome = factor(outcome, levels = outcomes)
  )
```

```{r descriptive-plots}
#| eval: false
# TPACK
p_tpack <- plot_outcome(
  data_long %>% filter(!str_detect(outcome, "^(AILIT|PU)")),
  scales_facet = "fixed"
) +
  ggtitle("Intelligent-TPACK")

ggsave(
  p_tpack + facet_wrap(~outcome, ncol = 3),
  file = "figures/plot-tpack.pdf",
  width = 8,
  height = 10,
  device = cairo_pdf
)
# AILIT and PU
p_ailit <- plot_outcome(
  data_long %>% filter(str_detect(outcome, "^(AILIT)"))
) +
  ggtitle("AI Literacy")

p_pu <- plot_outcome(
  data_long %>% filter(str_detect(outcome, "^(PU)")),
  scales_facet = "fixed"
) +
  ggtitle("Perceived Usefulness")
```
