---
title: "WoLKE-Studie-2026"
author: "Benedikt Beuttler"
date: "`r Sys.Date()`"
lang: en
page-layout: full
document-css: false
link-citations: true
number-sections: true
toc: true
csl: https://www.zotero.org/styles/apa
bibliography: library.bib
editor: source
execute:
  cache: false
format:
  html:
    standalone: true
    embed-resources: true
    section-divs: true
    html-math-method: mathjax
    wrap: none
    default-image-extension: png
    date-format: long  
    code-fold: true
    code-tools: true
  docx: 
    toc: true
    toc-depth: 2
    number-sections: true
    fig-caption: true
    fig-format: pdf
    default-image-extension: pdf
    date-format: long
    keep-md: true
    echo: false
    code-fold: true
prefer-html: true
---


```{r Information for first time user to get started with renv}
#| output: false
if (FALSE) {
  install.packages("renv") #1
  renv::restore() #2
}
```

# Preamble

## Prepare the Environment

## Load Libraries

```{r load-libraries-and-functions, cache=F}
#| output: false

# install pacman
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}
library(pacman)

## Pfad/Utilities
p_load(this.path)

## Wrangling
# use pacman to load packages
p_load(dplyr, tidyr, stringr, forcats, purrr, magrittr)

## Ähnlichkeit/Text
p_load(stringdist, psych)

## Modellierung
p_load(lavaan, lme4, lmerTest, emmeans, estimatr)

## Tabellen/Reporting
# install.packages("remotes")

# remotes::install_version(
#   "modelsummary",
#   version = "2.5.0",
#   repos = "https://cran.r-project.org"
# )

p_load(modelsummary)
p_load(kableExtra, arsenal)

## Visualisierung/Themes
p_load(ggplot2, ggsci, jtools)

p_load(mnormt)

source(file = "functions.r") # all outsourced funtions
```

```{r prepare-output}

is_word <- !(knitr::is_html_output() | knitr::is_latex_output())

if (is_word) {
  p_load(flextable)
}

```


## Load Pre- and Posttest and Reshape Data

```{r analysis-setup}
#define vars
controls <- c("gender", "semester", "location")
random_terms_lmer <- c("(1 | code)")
groups <- c("intervention")
time <- "time"
outcomes <- c(
  "AILIT",
  "AILIThorn",
  "AILITglat",
  "TK",
  "TPK",
  "TCK",
  "TPACKce",
  "TPACKgenai",
  "Ethics",
  "CEA",
  "CEA_understanding",
  "CEA_ped"
)
vars_num <- c("age", "semester") # numeric outcomes
vars_cat <- c("gender", "students") # categorical outcomes`
```

```{r load and prepare data}
prepare_data <- FALSE
if (prepare_data) {
  # Set working directory to the script's location
  # safe working-directory selection (replace setwd(this.path::this.dir()))
  wd <- tryCatch(this.path::this.dir(), error = function(e) NULL)

  if (is.null(wd)) {
    wd <- tryCatch(
      {
        if (
          requireNamespace("rstudioapi", quietly = TRUE) &&
            rstudioapi::isAvailable()
        ) {
          dirname(rstudioapi::getActiveDocumentContext()$path)
        } else if (requireNamespace("here", quietly = TRUE)) {
          here::here()
        } else {
          getwd()
        }
      },
      error = function(e) getwd()
    )
  }
  setwd(normalizePath(wd))

  # execute load_prepare_data.r to load and prepare the data via source command
  suppressMessages(source("./data/load_prepare_data.r"))
} else {
  # load pre-prepared data
  load("data/26-data_all.Rdata") # nicht nach pre und posttest gefiltert
  load("data/26-data.Rdata") # probanden mit pre UND posttest
  # load all Rdata from folder /data/dropouts
  files <- list.files("data/dropouts", pattern = "\\.Rdata$", full.names = TRUE)
  invisible(lapply(files, load, envir = .GlobalEnv))
}

data_pre <- data %>% filter(time == "pre")

```

```{r Prepare scales}

### grouping scales
## TPACK
# all TPACK items used
list_items_i_tpack <- list(
  "TK" = grep("^TP01_", names(data), value = T),
  "TPK" = grep("^TP02_", names(data), value = T),
  "TCK" = grep("^TP03_", names(data), value = T),
  "TPACK_ce" = grep("^TP04_(0[1-8]|12|15)$", names(data), value = T), # Sub-Subskala 1: Celik Enhanced Items
  "TPACK_genai" = grep("^TP04_(09|10|11|13|14)$", names(data), value = T), # Sub-Subskala 2: GenAI Items
  "Ethics" = grep("^TP05_0[1-6]$", names(data), value = T) #im pretest nur items 1-6 vorhanden 7,8 kamen im post dazu
)

# ein paar ethik items müssen noch rausgenommen werden, die nur im posttest erhoben wurden. hier nur check
# data_subset1 <- data %>% select(all_of(unlist(list_items_i_tpack)))
# View(data_subset1)

## Critical Evaluation of AI
list_items_cea <- list(
  "CEA" = grep("^RE02_(0[1-9])$", names(data), value = T),
  "CEA_understanding" = grep("^RE02_(0[1-4])$", names(data), value = T),
  "CEA_ped" = grep("^RE02_(0[5-9])$", names(data), value = T)
)
# data_subset3 <- data %>% select(all_of(unlist(list_items_cea)))
# sum(data_subset3 < 0, na.rm = TRUE)

## AI Literacy
# data <- data %>% mutate(AILIT = AI35_02) # rename to sumscore
list_items_ailit <- list(
  "ailit" = grep(
    "^(AI36_04|AI36_10|AI36_11|AI36_13|AI36_15|AI36_16|AI36_17|AI36_19|AI36_24|AI36_25|AI36_27|GL21_01|GL21_02|GL21_03|GL21_04|GL21_05|GL21_06|GL21_07|GL21_08|GL21_09|GL21_10|GL21_11|GL21_12|GL21_13|GL21_14|GL21_15|GL21_16|GL21_17|GL21_18|GL21_19)", # results of all ailit items
    names(data),
    value = T
  ),
  "ailit_horn" = grep(
    "^(AI36_04|AI36_10|AI36_11|AI36_13|AI36_15|AI36_16|AI36_17|AI36_19|AI36_24|AI36_25|AI36_27)",
    names(data),
    value = T
  ), # Hornberger items
  "ailit_glat" = grep(
    "^(GL21_01|GL21_02|GL21_03|GL21_04|GL21_05|GL21_06|GL21_07|GL21_08|GL21_09|GL21_10|GL21_11|GL21_12|GL21_13|GL21_14|GL21_15|GL21_16|GL21_17|GL21_18|GL21_19)",
    names(data),
    value = T
  ) # GLAT items
)

# create new columns with the means of the lists
data <- data %>%
  mutate(
    AILIT = rowSums(cbind(AI35_02, GL20_01), na.rm = TRUE), # default sumscore overall AI Lit
    AILIThorn = rowSums(
      select(., all_of(list_items_ailit$ailit_horn)),
      na.rm = TRUE
    ),
    AILITglat = rowSums(
      select(., all_of(list_items_ailit$ailit_glat)),
      na.rm = TRUE
    ),
    TK = rowMeans(select(., all_of(list_items_i_tpack$TK)), na.rm = TRUE),
    TPK = rowMeans(select(., all_of(list_items_i_tpack$TPK)), na.rm = TRUE),
    TCK = rowMeans(select(., all_of(list_items_i_tpack$TCK)), na.rm = TRUE),
    TPACKce = rowMeans(
      select(., all_of(list_items_i_tpack$TPACK_ce)),
      na.rm = TRUE
    ),
    TPACKgenai = rowMeans(
      select(., all_of(list_items_i_tpack$TPACK_genai)),
      na.rm = TRUE
    ),
    Ethics = rowMeans(
      select(., all_of(list_items_i_tpack$Ethics)),
      na.rm = TRUE
    ),
    CEA = rowMeans(select(., all_of(list_items_cea$CEA)), na.rm = TRUE),
    CEA_understanding = rowMeans(
      select(., all_of(list_items_cea$CEA_understanding)),
      na.rm = TRUE
    ),
    CEA_ped = rowMeans(select(., all_of(list_items_cea$CEA_ped)), na.rm = TRUE),
    TPACK_total = rowMeans(
      select(., all_of(unlist(list_items_i_tpack))),
      na.rm = TRUE
    )
  )

# check missing values in TPACK and PU scales (NA means, not answered (-1 diff. answer, -9 not answered))
data %>%
  select(all_of(c(list_items_i_tpack) %>% unlist)) %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  sum # 0

```

```{r reshape data}

# reshape to long
data_long <- data %>%
  pivot_longer(
    cols = c(all_of(outcomes)),
    names_to = "outcome",
    values_to = "score"
  )

data_long <- data_long %>%
  mutate(
    outcome = factor(outcome, levels = outcomes)
  )

```

```{r descriptive-plots}
# TPACK
p_tpack <- plot_outcome(
  data_long %>% filter(!str_detect(outcome, "^(AILIT|CEA)")),
  scales_facet = "fixed"
) +
  ggtitle("Intelligent-TPACK")

ggsave(
  p_tpack + facet_wrap(~outcome, ncol = 3),
  file = "figures/plot-tpack.pdf",
  width = 8,
  height = 10,
  device = cairo_pdf
)
# AILIT
p_ailit <- plot_outcome(
  data_long %>% filter(str_detect(outcome, "^(AILIT)"))
) +
  ggtitle("AI Literacy")

# Critical Evaluation of AI
p_cea <- plot_outcome(
  data_long %>% filter(str_detect(outcome, "^(CEA)"))
) +
  ggtitle("Critical Evaluation of AI (CEA)")

```


# Methods

## Participants

```{r participants, cache=F}
p_load(dplyr, tidyr) # needs to be called for rendering to work

data_participants <- data %>%
  filter(time == "pre") %>%
  group_by(intervention, location, gender) %>%
  summarise(
    n = n(),
    n_complete = sum(complete.cases(.)),
    age_min = min(age, na.rm = T),
    age_max = max(age, na.rm = T),
    age_mean = mean(age, na.rm = T) %>% round(2),
    age_sd = sd(age, na.rm = T) %>% round(2),
    semester_min = min(semester, na.rm = T),
    semester_max = max(semester, na.rm = T),
    semester_mean = mean(semester, na.rm = T) %>% round(2),
    semester_sd = sd(semester, na.rm = T) %>% round(2),
    .groups = "drop"
  )
```


```{r table-participants, result = "asis"}
#| label: tbl-participants
#| tbl-cap: "Demographic Data of the Three Experimental Groups."
# render as table
format <- ifelse(
  knitr::is_html_output(),
  "markdown",
  ifelse(knitr::is_latex_output(), "latex", "markdown")
)

var_names <- list(
  gender = "Gender",
  age = "Age",
  semester = "Semester Number",
  students = "Study Level",
  isp = "Completed Integrated Semester Internship",
  subject1 = "Subject 1",
  subject2 = "Subject 2",
  studies = "Study Program",
  location = "University",
  TK = "Intelligent TK",
  TCK = "Intelligent TCK",
  TPK = "Intelligent TPK",
  TPACK_total = "Intelligent TPACK",
  AILIT = "AI Literacy",
  Ethics = "AI Ethics",
  CEA = "Critical Evaluation of AI"
)

kable(
  summary(
    tableby(
      intervention ~
        location +
        age +
        gender +
        semester +
        students +
        TK +
        TCK +
        TPK +
        TPACK_total +
        Ethics +
        AILIT +
        CEA,
      data %>%
        filter(time == "pre") %>%
        mutate(
          students = factor(students, labels = c("Bachelor", "Master"))
        ) %>%
        droplevels(),
      cat.test = "chisq",
      chisq.correct = T,
      control = tableby.control(test = T, total = T, digits = 2)
    ),
    labelTranslations = var_names,
    pfootnote = T,
    text = F
  ),
  escape = T,
  format = format,
  booktabs = T
)
```

## Materials

### Data Preparation and Cleaning (needs to be done after data gets merged)

```{r}
# get participants who only participated in pre or post test
data_dropouts <- data_all %>%
  group_by(code) %>%
  summarise(
    has_pre = any(time == "pre"),
    has_post = any(time == "post"),
    .groups = "drop"
  ) %>%
  summarise(
    n_pre_only = sum(has_pre & !has_post),
    n_post_only = sum(!has_pre & has_post),
    n_both = sum(has_pre & has_post)
  )
```

## Statistical Analysis Ancova

```{r ancova}
# reshape data to wide format for ancova
data_wide <- data %>%
  pivot_wider(
    id_cols = c(
      code,
      intervention,
      group,
      gender,
      semester,
      students,
      location
    ),
    names_from = time,
    values_from = all_of(outcomes)
  ) %>%
  mutate(
    # scale the posttest score so that the estimates are approximations of Cohen's d
    across(all_of(outcomes %>% paste0("_post")), scale),
    treatment = factor(ifelse(
      intervention == "control",
      "control",
      "intervention"
    ))
  )

contrasts.treatment <- cbind(c(-1 / 2, 1 / 2))
colnames(contrasts.treatment) <- c("-c+t")
contrasts(data_wide$treatment) <- contrasts.treatment

ancovas <- lapply(
  outcomes,
  fit_ancova,
  groups,
  # we remove University as a covariate as we only have control students at University 2
  setdiff(controls, "location"),
  "HC3",
  data_wide
)
names(ancovas) <- outcomes

ancovas_two_groups <- lapply(
  outcomes,
  fit_ancova,
  "treatment",
  setdiff(controls, "location"),
  "HC3",
  data_wide
)
names(ancovas_two_groups) <- outcomes

```


## AI Literacy

```{r fig-ai-lit, echo=FALSE, fig.cap="Descriptive Statistics of Students' AI Literacy Scores Over Time by Treatment Group.", out.width="100%",fig.width=7, fig.height=6.5}
plot_outcome(
  data_long %>%
    filter(outcome %in% c("AILIT", "AILIThorn", "AILITglat")) %>%
    mutate(
      outcome = factor(
        outcome,
        levels = c("AILIT", "AILIThorn", "AILITglat"),
        labels = c(
          "AI Literacy - Full Scale",
          "AI Literacy - Study-Specific",
          "AI Literacy - GLAT"
        )
      )
    ),
  ylab = "Score"
) +
  ggtitle("AI Literacy") +
  theme(legend.position = "bottom")

#p_ailit %+% + facet_wrap(~outcome, ncol =2, scales="free")  + labs(y = "Score")  + theme(legend.position = "bottom")

```

```{r tab.layout = "autofit"}
#| label: tbl-ancova-ailit-summary
#| tbl-cap: "Summary of Multiple Regression Models for AI Literacy."

ft_ailit_ancova <- modelsummary_models(
  ancovas[names(ancovas) %in% c("AILIT", "AILIThorn", "AILITglat")],
  outcomes = c("AILIT", "AILIThorn", "AILITglat"),
  output = modelsummary_output
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_ailit_ancova <- adjust_table_for_word(ft_ailit_ancova, width = .75)
}
ft_ailit_ancova
```

AILIT_post ~ intervention + pretest + gender + semester

interventionstem = Unterschied zwischen STEM und CONTROL

interventionlanguage = Unterschied zwischen Language und CONTROL

Das Modell testet nicht direkt STEM vs Language.

b interessiert und hier und ob der Wert sign (*) ist oder nicht.

| Outcome   | STEM Effekt | Language Effekt |
| --------- | ----------- | --------------- |
| AILIT     | **0.92***   | **0.56***       |
| AILIThorn | **0.49***   | n.s.            |
| AILITglat | **1.05***   | **0.61**        |

```{r, results="asis"}
if (!is_word) {
  cat("### Pairwise Contrasts Between Groups")
}
```

```{r ailit-ancova-between-contrasts}
# pairwise between contrasts (no multiplicity adjustment)
ailit_ancova_between_contrasts <- lapply(ancovas[1:3], function(m) {
  emmeans(m, ~intervention) %>%
    contrast(interaction = "pairwise", adjust = "none")
})
if (!is_word) {
  print(ailit_ancova_between_contrasts)
}
```


## Intelligent TPACK and AI Ethics

```{r fig-tpack, echo=FALSE, fig.cap="Descriptive Statistics of Students' Intelligent TPACK and AI Ethics Ratings Over Time by Treatment Group.", out.width="100%",fig.width=7, fig.height=6.5}
p_tpack +
  facet_wrap(
    ~outcome,
    scales = "fixed",
    ncol = 3,
    labeller = labeller(
      outcome = c(
        TK = "Intelligent TK",
        TCK = "Intelligent TCK",
        TPK = "Intelligent TPK",
        TPACKce = "Intelligent TPACK - Core",
        TPACKgenai = "iTPACK - Generative AI",
        Ethics = "AI Ethics"
      )
    )
  ) +
  ggtitle("Intelligent TPACK and AI Ethics") +
  theme(legend.position = "bottom")
```


::: {.landscape}
```{r tab.layout="autofit"}
#| label: tbl-ancova-tpack-summary
#| tbl-cap: "Summary of Multiple Regression Models for Intelligent TPACK Subscales and AI Ethics"

ft_ancova <- modelsummary_models(
  ancovas[grepl("^(T|Ethics)", names(ancovas))],
  outcomes = outcomes %>% grep("^(T|Ethics)", ., value = TRUE),
  output = modelsummary_output,
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_ancova <- adjust_table_for_word(ft_ancova, width = 1)
}
ft_ancova
```

:::

## Critical Evaluation of AI (CEA)
Note für uns: ggf. reduzieren wir CEA auf eine Skala mit 9 Items. Im Piloten haben wir 15 items getestet und EFA hat gezeigt, dass Reduktion auf 2 Skalen möglich wäre. Das haben wir gemacht, und die funktionieren auch, aber ein Item lädt z.B. ziemlich stark auf die andere Skala. Entweder bekommt man das gut begründet, aber den richtigen Mehrwert sehe ich nicht, 2 Skalen aufzustellen. So oder so, es haben alle Skalen gut funktioniert und zeigen den Mehrwert vom Seminar.

```{r fig-cea, echo=FALSE, fig.cap="Descriptive Statistics of Students' Critical Evaluation of AI Over Time by Treatment Group.", out.width="100%",fig.width=7, fig.height=6.5}
p_cea +
  facet_wrap(
    ~outcome,
    scales = "fixed",
    ncol = 2,
    labeller = labeller(
      outcome = c(
        CEA = "Critical Evaluation of AI (CEA) - Full Scale",
        CEA_understanding = "Critical Understanding of AI",
        CEA_ped = "Pedagogical Evaluation of AI Output"
      )
    )
  ) +
  ggtitle("Critical Evaluation of AI (CEA)") +
  theme(legend.position = "bottom")
```

::: {.landscape}
```{r tab.layout="autofit"}
#| label: tbl-ancova-CEA-summary
#| tbl-cap: "Summary of Multiple Regression Models for CEA subscale."

ft_cea_ancova <- modelsummary_models(
  ancovas[grepl("^(CEA)", names(ancovas))],
  outcomes = c("CEA", "CEA_understanding", "CEA_ped"),
  output = modelsummary_output,
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_cea_ancova <- adjust_table_for_word(ft_cea_ancova, width = .75)
}
ft_cea_ancova
```

:::

``` {r Exploring CEA data}
# cronbach alpha
alphas <- imap(
  list_items_cea,
  ~ psych::alpha(dplyr::select(data_pre, dplyr::all_of(.x)), warnings = FALSE)
)
alphas
# 0.86 alpha, kein item lohnt sich rauszunehmen

data_post <- data %>% filter(time == "post" & CEA != "NaN")
# EFA
ncomp <- 2
# EFA
df_cea_items <- data_post %>%
  select(all_of(list_items_cea$CEA)) %>%
  na.omit()
cea_scale <- list_items_cea %>% unlist(use.names = F)

efa_promax <- df_cea_items %>%
  select(any_of(cea_scale)) %>%
  psych::fa(nfactors = ncomp, rotate = "varimax", fm = "ml", missing = T)

# print(efa_promax, digits = 2, cut = .3, sort = F)
fa.diagram(efa_promax, digits = 2, cut = .30, simple = FALSE)

## Computing EFA using the lavaan package
cea_efa <- lavaan::efa(
  df_cea_items %>%
    select(any_of(cea_scale)),
  nfactors = ncomp,
  rotation = "varimax",
  ordered = F,
  std.lv = TRUE,
  missing = "fiml"
)

summary(cea_efa, nd = 3L, cutoff = 0.3, dot.cutoff = 0.05)

```

Results from efa and cronbachs alpha confirm our exploration and adjustments from pilot data. we have a good scale in CEA.

## Two Groups (Control vs. Intervention)

```{r ancova-two-groups}
#| label: tbl-ancova-two-groups
#| tbl-cap: "Summary of Multiple Linear Regression Models for Two Treatment Groups."

modelsummary_models(
  ancovas_two_groups,
  outcomes = outcomes,
  output = modelsummary_output
)

```


```{r save-env }
# save environment
save.image("data/26-env-analysis.Rdata")
```