---
title: "WoLKE CAEO 2025 Report"
author: "Heiko Holz"
date: "`r Sys.Date()`"
lang: en
page-layout: full
document-css: false
link-citations: true
number-sections: true
toc: true
csl: https://www.zotero.org/styles/apa
bibliography: library.bib
editor: source
execute:
  cache: true
format:
  html:
    standalone: true
    embed-resources: true
    section-divs: true
    html-math-method: mathjax
    wrap: none
    default-image-extension: png
    date-format: long  
    code-fold: true
    code-tools: true
  docx: 
    toc: true
    toc-depth: 2
    number-sections: true
    fig-caption: true
    fig-format: pdf
    default-image-extension: pdf
    date-format: long
    keep-md: true
    echo: false
    code-fold: true
prefer-html: true
---

This document contains the analysis scripts and report for the manuscript "A Course on AI for Pre-Service STEM and Language Teachers: A Quasi-Experimental Evaluation of Learning Outcomes Using Intelligent-TPACK and AI Literacy Instruments ". The R scripts, data are hosted at <OSF>.

```{r load-libraries-and-functions, cache=F}
#| output: false
library(this.path)
library(dplyr) #e.g. for select
library(tidyr) # for reshaping data
library(stringdist) #levenshtein
library(stringr) # for reshaping data
library(lme4)
library(lmerTest)
library(emmeans)
library(modelsummary)
library(kableExtra)
library(arsenal)
library(forcats)
library(ggplot2)
library(ggsci) # color palettes
library(jtools) # theme_apa etc.
library(papaja) # for apa reporting
library(sjPlot) # for tab_model
library(stargazer) # for stargazer tables
library(estimatr) # for lm_robust
source(file = "functions.r") # all outsourced funtions
```


```{r load-data}
# load "env-analysis.RData" which contains all data needed for the analysis
load(file = "data/env-analysis.RData")

names(lmers_prepost_all) <- outcomes
# get the number of the terms in the model for the two treatment contrasts
tc2 <- nrow(summary(lmers_prepost_all[[1]])$coefficients)
tc1 <- tc2 - 1
```

```{r rerun-models-with-proper-contrasts}
data_wtf <- data %>%
  droplevels %>%
  mutate(
    location = factor(
      location,
      levels = c("PH Schwäbisch Gmünd", "PH Ludwigsburg")
    ),
    treatment = factor(ifelse(intervention == "control", "control", "intervention"))
  )

## define contrast Control vs Exp1+Exp2 - vermutlich bessere version, da die Estimates "stimmen"
mat <- cbind(
  c(1 / 3, 1 / 3, 1 / 3), # Durchschnitt aller 3 Gruppen
  c(-1, .5, .5), # Vergleich Control vs Mittel der beiden Interventions-Gruppen
  c(0, -1, 1)) # Vergleich Sprache vs MINT
contrasts.CvsExp <- solve(t(mat))[, -1]
colnames(contrasts.CvsExp) <- c("-c+STEM_language", "-STEM+language")
contrasts(data_wtf$intervention) <- contrasts.CvsExp

# GENDER
# female: -1/2
# male : +1/2
contrasts.gender <- cbind(c(-1 / 2, 1 / 2))
colnames(contrasts.gender) <- c("-m+f")
contrasts(data_wtf$gender) <- contrasts.gender

# LOCATION
# SG UE: -1/2
# L UE: +1/2
contrasts.location <- cbind(c(-1 / 2, 1 / 2))
colnames(contrasts.location) <- c("-SGUE+LUE")
contrasts(data_wtf$location) <- contrasts.location

# TIME
# pre: -1/2
# post: +1/2
contrasts.time <- cbind(c(-1 / 2, 1 / 2))
colnames(contrasts.time) <- c("-pre+post")
contrasts(data_wtf$time) <- contrasts.time

data <- data_wtf
# refit all lmer_prepost_all with new data
lmers_prepost_all <- lapply(
  outcomes,
  fit_lmer,
  groups,
  time,
  controls,
  random_terms_lmer,
  NULL,
  data = data
)
names(lmers_prepost_all) <- outcomes

```

```{r prepare-output}

is_word <- !(knitr::is_html_output() | knitr::is_latex_output())

if (is_word) {
  library(flextable)
}

modelsummary_output <- ifelse(
  knitr::is_latex_output(),
  "latex_tabular",
  ifelse(knitr::is_html_output(), "kableExtra", "flextable")
)

adjust_table_for_word <- function(ft, width = 0) {
  ft <- flextable::autofit(ft)
  ft <- flextable::set_table_properties(ft, layout = "autofit", width = width)
  ft <- flextable::fontsize(ft, size = 10, part = "header")
  ft <- flextable::fontsize(ft, size = 8, part = "body")
  ft
}
```

# Methods

## Participants

```{r}
data_participants <- data %>%
  filter(time == "pre") %>%
  group_by(intervention, location, gender) %>%
  summarise(
    n = n(),
    n_complete = sum(complete.cases(.)),
    age_min = min(age, na.rm = T),
    age_max = max(age, na.rm = T),
    age_mean = mean(age, na.rm = T) %>% round(2),
    age_sd = sd(age, na.rm = T) %>% round(2),
    semester_min = min(semester, na.rm = T),
    semester_max = max(semester, na.rm = T),
    semester_mean = mean(semester, na.rm = T) %>% round(2),
    semester_sd = sd(semester, na.rm = T) %>% round(2)
  ) %>%
  ungroup()
```


The study involved pre-service teachers from two universities of education in Baden-Württemberg, Germany. Participants were divided into a **treatment group** consisting of two intervention groups (STEM and language course), and a **control group**.
Students in the treatment groups self-selected the courses "AI in STEM/Language Education" as part of their compulsory elective studies. The control group was recruited from other STEM courses at University 1 during the same semester. 

In total, $N=`r data_all %>% filter(time=="pre") %>% nrow()`$ students originally participated in the study ($N_c=`r data_all %>% filter(time=="pre", intervention=="control") %>%nrow`$ in the control group, $N_s= `r data_all %>% filter(time=="pre", intervention=="stem") %>%nrow `$ in the STEM course, and $N_l=`r data_all %>% filter(time=="pre", intervention=="language") %>%nrow `$ in the language course), of which $n=`r data_participants$n %>% sum`$ completed both pre- and the posttest and were included in the analyses ($n_c = `r data_participants%>%filter(intervention=="control") %>% select(n)%>%sum`$, $n_s = `r data_participants%>%filter(intervention=="stem") %>% select(n)%>%sum`$, $n_l = `r data_participants%>%filter(intervention=="language") %>% select(n) %>% sum`$).

Unfortunately, we observed a high drop-out rate between the pre- and posttest. We attribute this to several factors: First, the posttests were conducted in class close  (one to two weeks) to the examination period, resulting in students not attending the in-person posttest sessions. For the control group, some lecturers executed the posttest during their teaching, while others did not. We suspect a high drop-out at the posttest for the control group due to both factors: students not attending the sessions close to the examination week and not conducting the tests at home or in their free study time.


Descriptive statistics of students' demographics and pretest scores are presented in [@tbl-participants]. The final sample size included in the analyses was $n=76$, aged between $`r data_pre$age%>%min`$ and $`r data_pre$age %>% max`$ years ($M=`r data_pre$age%>%mean %>% round(2)`$, $SD=`r data_pre$age%>%sd%>%round(2)`$), of which $`r data_pre%>%filter(gender=="Weiblich")%>%nrow`$ identified as female and $`r data_pre%>%filter(gender=="Männlich")%>%nrow`$ as male.
The students were enrolled in semester $`r data_pre$semester%>%min`$ to $`r data_pre$semester%>%max`$ ($M=`r data_pre$semester%>%mean%>%round(2)`$, $SD=`r data_pre$semester%>%sd%>%round(2)`$). Of these, $`r data_pre%>%filter(students=="bachelor")%>%nrow`$ students were at the Bachelor's level and $`r data_pre%>%filter(students=="master")%>%nrow`$ at the Master's level.


```{r table-participants, result = "asis"}
#| label: tbl-participants
#| tbl-cap: "Demographic Data of the Three Experimental Groups."
# render as table
format <- ifelse(
  knitr::is_html_output(),
  "markdown",
  ifelse(knitr::is_latex_output(), "latex", "markdown")
)

var_names <- list(
  gender = "Gender",
  age = "Age",
  semester = "Semester Number",
  students = "Study Level",
  isp = "Completed Integrated Semester Internship",
  subject1 = "Subject 1",
  subject2 = "Subject 2",
  studies = "Study Program",
  location = "University",
  TK = "Intelligent TK",
  TCK = "Intelligent TCK",
  TPK = "Intelligent TPK",
  TPACK_total = "Intelligent TPACK",
  AILIT = "AI Literacy",
  PU = "Perceived Usefulness",
  Ethics = "AI Ethics"
)

kable(
  summary(
    tableby(
      intervention ~
        location +
          age +
          gender +
          semester +
          students +
          TK +
          TCK +
          TPK +
          TPACK_total +
          Ethics +
          AILIT +
          PU,
      data %>%
        filter(time == "pre") %>%
        mutate(
          gender = factor(gender, labels = c("Female", "Male")),
          students = factor(students, labels = c("Bachelor", "Master"))
        ) %>%
        droplevels(),
      cat.test = "chisq",
      chisq.correct = T,
      control = tableby.control(test = T, total = T, digits = 2)
    ),
    labelTranslations = var_names,
    pfootnote = T,
    text = F
  ),
  escape = T,
  format = format,
  booktabs = T
)
```

### Effect Sizes for Group Differences for Covariates
```{r group-differences-effectsizes}

vars_num <- c("age", "semester")      # numeric outcomes
vars_cat <- c("gender", "students")   # categorical outcomes

d <- data %>%
  filter(time == "pre") %>%
  mutate(
    gender   = factor(gender,   labels = c("Female", "Male")),
    students = factor(students, labels = c("Bachelor", "Master"))
  )

# eta^2 for numeric outcomes
eta_num <- purrr::map_dfr(vars_num, function(v) {
  f <- as.formula(paste(v, "~ intervention"))
  m <- lm(f, data = d)
  es <- effectsize::eta_squared(m, partial = TRUE)
  tibble(
    variable = v,
    effect   = "eta2",
    value    = es$Eta2[1],
    interpretation = effectsize::interpret_eta_squared(value) %>% as.character
  )
})

# Cramér's V for categorical outcomes
cv_cat <- purrr::map_dfr(vars_cat, function(v) {
  tab <- table(d[[v]], d$intervention)
  cv  <- effectsize::cramers_v(tab)
  tibble(
    variable = v,
    effect   = "cramers_v",
    value    = cv$Cramers_v,
    interpretation = effectsize::interpret_cramers_v(value) %>% as.character
  )
})

effectsizes_covariates <- bind_rows(eta_num, cv_cat) %>%
  mutate(
    value = apa_num(value)
  )
effectsizes_covariates %>% kable

cohens_d_two_groups <- purrr::map_dfr(vars_num, function(v) {
  es <- effectsize::cohens_d(v, "treatment", d, pooled = TRUE)
  tibble(
    variable = v,
    effect   = "cohens_d",
    value    = es$Cohens_d,
    interpretation = effectsize::interpret_cohens_d(value) %>% as.character
  )
})

phi_two_groups <- purrr::map_dfr(vars_cat, function(v) {
  tab <- table(d[[v]], d$treatment)
  phi <- effectsize::phi(tab, adjust = F)
  tibble(
    variable = v,
    effect   = "phi",
    value    = phi$phi,
    interpretation = effectsize::interpret_phi(value) %>% as.character
  )
})

# combine
effectsizes_covariates__two_groups <- bind_rows(cohens_d_two_groups, phi_two_groups) %>%
  mutate(
    value = apa_num(value)
  )
```

## Materials

### Translation Process
The Intelligent-TPACK, originally developed by @celik2023, is widely used to assess teachers’ knowledge for pedagogical and ethical AI use based on the technological, pedagogical, and content knowledge (TPACK) framework.
However, no empirically validated German translations of the Intelligent-TPACK exists. While some studies translate and use subsets of the Intelligent-TPACK [e.g., @mah.gross2024], there is no validated German version of the full instrument. 

Therefore, for the Intelligent-TPACK by @celik2023 as well as for items from the AI ethics subscale from @chiu.ahmad.ea2024, we translated the original items in a participative multi-step process involving not just experts from the fields of computer science and education, but also (pre-service) teachers.
For this, we applied the *Modified Direct Translation* that includes translators and experts discussing the wording until reaching agreement [@behling.law2025]: First, seven researchers and experts in AI ED (additionally including one English teacher and one student teacher of Computer Science and English) translated the items independently. Second, we compared all translations and marked the items that received different translations. In a third step, we performed a group workshop with all translators in which the marked items were discussed and dot voting was applied to decide on the most apporopriate translation. Finally, three items that were still marked as questionable were discussed and further adapted with a native English teacher who teaches at University 2.

The translation and validation of the Intelligent-TPACK is reported in detail in @authorsinprep.

## Analysis
We analyzed the interventions' efficacy in a multilevel analysis approach. To contrast the differences of students' change in the respective outcome measure over time, we implemented a linear mixed-effects regression model for each outcome measure, considering our $3$ factorial design. To account for the panel structure, that is, two test sessions, in the data, we used the random effects (RE) estimation method to estimate treatment effects, following CITATION.

The raw data and *R* scripts detailing the analyses can be accessed in the online supplement at <OSF>.

:::{.callout-note}
Ist unser Design 3- oder 2-faktoriell?
-> wenn treatment 3 Gruppen, dann 3-faktoriell, sonst 2-faktoriell
:::

We implemented the two-level models with the students at level 1 (individual level) and the University at level 2 (class level). We addressed the clustering of time points within students by including random intercepts for students. The clustering of students within universities was addressed by including a dummy variable for the University. 

To contrast the differences in change over time between the treatment groups (i.e., estimated treatment effect), we included the time-by-treatment interaction in the models. For this, we used a dummy-coded *treatment* variable using reversed Helmert contrasts to contrast the differences between (i) the *control* group and the *intervention* groups and (ii) between the two intervention groups. We set the control group as the reference group.

Finally, we included variables as a covariate if they met the criteria for baseline equivalence in randomized controlled trials by the What Works Clearinghouse [@WWC2022]. That is, we included the variables as covariates if the difference between the groups was statistically significant, or if it exceeded $d>.05$.
We tested for group differences concerning the following variables at baseline (pretest): age, gender, semester, and study level (Bachelor vs. Master). We found for gender and semesters  statistically significant differences between the treamtment groups, see [@tbl-participants]. We thus included the following covariates in the models: gender and semester.

That is, the final model for each outcome measure contained time, treatment, gender, semester, University and the interaction of time and treatment as fixed effects and a random intercept for students, and was specified as follows:
`lmer(outcome ~ time * treatment + gender + semester + university + (1|student))`.

The criterion of statistical significance was set at α=.05.

### Data Preparation and Cleaning


```{r}
# get participants who only participated in pre or post test
data_dropouts <- data_all %>%
  group_by(code) %>%
  summarise(
    has_pre = any(time == "pre"),
    has_post = any(time == "post"),
    .groups = "drop"
  ) %>%
  summarise(
    n_pre_only = sum(has_pre & !has_post),
    n_post_only = sum(!has_pre & has_post),
    n_both = sum(has_pre & has_post)
  )
```

For data cleaning, we applied the following steps: First, we investigated which students, based on their pseudonyms, did not participate in both test sessions. Based on a string distance analysis using the Levenshtein distance [@levenshtein1966], we identified pseudonyms that were likely to belong to the same student but were entered slightly differently (e.g., due to typos or students who moved during the semester, as the address was part of their pseudonym). We then manually checked these pseudonyms and merged them if we believe they belonged to the same student. This resulted in $n=24$ students being merged.
We then removed students who only participated at the pretest ($n=`r data_dropouts$n_pre_only`$) or only at the posttest ($n=`r data_dropouts$n_post_only`$).
Further, some students conducted the post test twice. In this case, we only considered their first response in the analyses and dropped the second.

As for some self-constructed items students could choose "I don't understand the item" as a response option, we recoded these responses as missing values ($n=`r data %>%
  select(all_of(c(list_items_i_tpack, list_items_pu)%>%unlist)) %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>% sum`$ responses to individual items).

For the Intelligent TPACK, AI ethics and perceived usefulness (sub)scales, we calculated the mean score across all items of the respective (sub)scale per student. For AI literacy, we calculated the sum score across all items. We additionally calculated the sum score for the subset of items of the AI literacy   that that directly reflect the topics addressed in the course, see SECTION MATERIALS.

### Statistical Software
We performed the statistical analyses in *R* [@R2024] using the following packages:
*dplyr* [wickham2023:dplyr], *tidyr* [@wickham2024], and *gtools* [@warnes2023] for data wrangling;
*arsenal* [@heinzen2021], *vtable* [@huntington2023], *modelsummary* [@arel-bundock2022] and *kableExtra* [@zhu2024] for creating tables; *ggplot2* [@wickham2016], *ggsci* [@xiao2024] and *jtools* [@long2022] for plotting; and *lme4* [@bates2015] and *lmerTest* [@kuznetsova2017] for linear mixed-effects regression analyses.

We used the *lme4* package [@bates2015] to fit linear mixed-effects models and the *broom.mixed* package [@bolker2024] to extract the results. Degrees of freedom and *p*-values were calculated using the *lmerTest* package [@kuznetsova2017] using Kenward-Roger's approximation for the denominator degrees of freedom. Pairwise comparisons between the treatment groups were computed using the *emmeans* package [@lenth2025].


# Results

Descriptive statistics of students' pretest scores are presented in [@tbl-participants].
As expected, students attending the STEM course and students from the control group, which was recruited from pre-service STEM teachers, reported significantly higher ratings in all Intelligent TPACK subscales compared to students attending language course, which predominantly consisted of students not studying a STEM subject as their second subject. We didn't find significant differences between the intervention groups for AI literacy, AI ethics, and perceived usefulness at the pretest.

The descriptive statistics of students' change in the outcome measures over time per treatment group are listed in  [@fig-ai-lit], [@fig-tpack], and [@fig-pu].

For all outcome measure, we will first report the *main* treatment effect, i.e., the interaction between time and treatment group. Then, we will report the differences between treatment groups, i.e., control vs. intervention groups (STEM and language course), and then contrast the intervention groups, i.e., STEM vs. language course).

The results of the linear mixed-effects models for each outcome measure are presented in [@tbl-lmer-ailit-summary], [@tbl-lmer-tpack-summary], and [@tbl-lmer-pu-summary].

## Effect Sizes

```{r effect-sizes}
dhat <- function(model, idx = 3, type = 2) {
  f <- abs(tail(summary(model)$coefficients, idx))[1]
  v <- sqrt(tail(anova(model, type = type)[["Mean Sq"]], 1))
  return(f / v)
}
#' Compute standardized effect size d for one fixed effect in an lmer model, based on https://pmc.ncbi.nlm.nih.gov/articles/PMC6646942/
#'
#' d = estimate / sqrt(sum of variances of random effects [+ residual if chosen])
#'
#' @param model lmer model
#' @param effect character, name of the fixed effect in fixef(model)
#' @param include_residual logical, include residual variance (default TRUE)
#' @return numeric, effect size d
lmer_d <- function(model, effect, include_residual = TRUE) {
  stopifnot(effect %in% names(lme4::fixef(model)))

  # numerator
  est <- unname(lme4::fixef(model)[effect])

  # variance components
  vc <- as.data.frame(lme4::VarCorr(model)) # grp, var1, var2, vcov, sdcor
  var_random <- sum(vc$vcov, na.rm = TRUE)

  # residual
  if (!include_residual) {
    var_random <- var_random - sigma(model)^2
  }

  d <- est / sqrt(var_random)
  return(d)
}

effect_sizes_dhat_old <- sapply(lmers_prepost_all, dhat, idx = 2) %>% round(2)

effect_sizes <- sapply(
  lmers_prepost_all,
  lmer_d,
  "intervention-c+STEM_language:time-pre+post"
) %>%
  round(2)
effect_sizes
```

## AI Literacy
Descriptive results of students' AI literacy scores are shown in [@fig-ai-lit], and the results of the linear mixed-effects models are listed in [@tbl-lmer-ailit-summary].

```{r fig-ai-lit, echo=FALSE, fig.cap="Descriptive Statistics of Students' AI Literacy Scores Over Time by Treatment Group.", out.width="80%",fig.width=6, fig.height=4.5}
plot_outcome(
  data_long %>%
    filter(outcome %in% c("AILIT", "AILITWolke")) %>%
    mutate(
      outcome = factor(
        outcome,
        levels = c("AILIT", "AILITWolke"),
        labels = c("AI Literacy - Full Scale", "AI Literacy - Study-Specific")
      )
    ),
  ylab = "Score"
) +
  ggtitle("AI Literacy") +
  theme(legend.position = "bottom")

#p_ailit %+% + facet_wrap(~outcome, ncol =2, scales="free")  + labs(y = "Score")  + theme(legend.position = "bottom")

# TODO: auch AI Lit Short für Online Supplement
```

```{r tab.layout="autofit"}
#| label: tbl-lmer-ailit-summary
#| tbl-cap: "Summary of Linear Mixed-Effects Models for AI Literacy."

ft_ailit <- modelsummary_models(
  lmers_prepost_all[names(lmers_prepost_all) %in% c("AILIT", "AILITWolke")],
  ddf = "kr",
  output = modelsummary_output,
  # coef_rename = c("genderMännlich" = "Geschlecht")
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_ailit <- adjust_table_for_word(ft_ailit, width = .75)
}
ft_ailit
```

### AI Literacy -- Full Scale
We did not find a significant main effect of treatment on students' change in AI literacy, `r report_lmer_anova(lmers_prepost_all[["AILIT"]])`. Neither did the control group's change differ significantly from both intervention groups, 
`r report_lmer(lmers_prepost_all[["AILIT"]], term = tc1)`, nor did the change differ significantly between the two intervention groups, `r report_lmer(lmers_prepost_all[["AILIT"]], term = tc2)`.

However, as [@fig-ai-lit] shows, backened by within-group contrast analyses contrasting the AI literacy score between pre- and posttest (see Online Supplement), students in all treatment conditions improved their AI literacy, while the change in pre-service STEM teachers was significant, `r report_contrast(ailit_within_contrasts[[1]], 2)`.

### AI Literacy -- Study-Specific Subset
For the study-specific subset of the AI literacy items, we also did not find a significant main effect of treatment on students' change in AI literacy, `r report_lmer_anova(lmers_prepost_all[["AILITWolke"]])`. The control group's change did not differ significantly from both intervention groups, 
`r report_lmer(lmers_prepost_all[["AILITWolke"]], term = tc1)`, nor did the change between the two intervention groups, `r report_lmer(lmers_prepost_all[["AILITWolke"]], term = tc2)`.

```{r, results="asis"}
if (!is_word) {
  cat("#### Pairwise Contrasts Within Groups")
}
```

```{r ailit-within-contrasts}

# pairwise within contrasts (no multiplicity adjustment)
ailit_within_contrasts <- lapply(lmers_prepost_all[1:3], function(m) {
  emmeans(m, ~ time | intervention) %>%
    contrast(method = "revpairwise", adjust = "none")
})
if (!is_word) {
  print(ailit_within_contrasts)
}
```

```{r, results="asis"}
if (!is_word) {
  cat("#### Pairwise Contrasts Between Groups")
}
```

```{r ailit-between-contrasts}
# pairwise between contrasts (no multiplicity adjustment)
ailit_between_contrasts <- lapply(lmers_prepost_all[1:3], function(m) {
  emmeans(m, ~ intervention * time) %>%
    contrast(interaction = "revpairwise", adjust = "none")
})
if (!is_word) {
  print(ailit_between_contrasts)
}
```

However, as can be seen in [@fig-ai-lit], students in both intervention groups showed descriptively higher increases in AI literacy compared to the control group. When performing post-hoc contrasts, we found that students in the STEM course had a marginally significantly higher increase in the study-specific AI literacy compared to the control group, `r report_contrast(ailit_between_contrasts[[2]], 1)`, while the difference between students in the language course and the control group was insignificant, `r report_contrast(ailit_between_contrasts[[2]],2)`. All contrast analyses can be accessed in the online supplement at <OSF>.


## Intelligent TPACK and AI Ethics
The descriptive results of students' ratings of perceived knowledge in Intelligent TPACK and AI ethics are shown in [@fig-tpack], and [@tbl-lmer-tpack-summary] lists the results of the linear mixed-effects models. As can be seen in [@fig-tpack], all pre-service teachers reported their intelligent TPACK knowledge to be around or below the middle of the scale at the pretest, while the intervention groups reported, on average, values significantly higher than the middle value of the scale at the posttest.

For all subscales related to Intelligent TPACK and AI ethics, that is, Intelligent TK, Intelligent TPK, Intelligent TCK, Intelligent TPACK, and AI ethics, we found a significant main effect of treatment on student' change on the respective subscales. We found significantly higher increases in students' Intelligent TPACK subscales in both intervention groups compared to the control group, whereas the intervention groups did not differ significantly from each other, see [@tbl-lmer-tpack-summary] for an overview.

```{r fig-tpack, echo=FALSE, fig.cap="Descriptive Statistics of Students' Intelligent TPACK and AI Ethics Ratings Over Time by Treatment Group.", out.width="100%",fig.width=7, fig.height=6.5}
p_tpack +
  facet_wrap(
    ~outcome,
    scales = "fixed",
    ncol = 3,
    labeller = labeller(
      outcome = c(
        TK = "Intelligent TK",
        TCK = "Intelligent TCK",
        TPK = "Intelligent TPK",
        TPACKce = "Intelligent TPACK - Core",
        TPACKgenai = "iTPACK - Generative AI",
        Ethics = "AI Ethics"
      )
    )
  ) +
  ggtitle("Intelligent TPACK and AI Ethics") +
  theme(legend.position = "bottom")
```

::: {.landscape}
```{r tab.layout="autofit"}
#| label: tbl-lmer-tpack-summary
#| tbl-cap: "Summary of Linear Mixed-Effects Models for Intelligent TPACK Subscales and AI Ethics"

ft <- modelsummary_models(
  lmers_prepost_all[grepl("^(T|Ethics)", names(lmers_prepost_all))],
  ddf = "kr",
  output = modelsummary_output,
  # coef_rename = c("genderMännlich" = "Geschlecht")
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft <- adjust_table_for_word(ft, width = 1)
}
ft
```

:::

### Intelligent TK
We found a significant main effect of treatment on students' change in intelligent TK, `r report_lmer_anova(lmers_prepost_all[["TK"]])`. 
We found a significant higher increase in students' intelligent TK in both intervention groups compared to the control group, `r report_lmer(lmers_prepost_all[["TK"]], term = tc1)`, while the change did not differ significantly between the two intervention groups, `r report_lmer(lmers_prepost_all[["TK"]], term = tc2)`.

### Intelligent TPK
We found a significant main effect of treatment on students' change in intelligent TCK, `r report_lmer_anova(lmers_prepost_all[["TPK"]])`.
We found a significant higher increase in students' intelligent TCK in both intervention groups compared to the control group, `r report_lmer(lmers_prepost_all[["TPK"]], term = tc1)`, whereas the two intervention groups did not differ significantly in their change, `r report_lmer(lmers_prepost_all[["TPK"]], term = tc2)`.

### Intelligent TCK
We found a significant main effect of treatment on students' change in intelligent TPK, `r report_lmer_anova(lmers_prepost_all[["TCK"]])`.
We found a significant higher increase in students' intelligent TPK in both intervention groups compared to the control group, `r report_lmer(lmers_prepost_all[["TCK"]], term = tc1)`, but the change did not differ significantly between the two intervention groups, `r report_lmer(lmers_prepost_all[["TCK"]], term = tc2)`.

### Intelligent TPACK
For both subscales of the Intelligent TPACK, that is, students' general TPACK and their perceived competence of applying *generative AI* in teaching their subject, we found a significant main effect of treatment on students' change on the respective subscales, *TPACK*: `r report_lmer_anova(lmers_prepost_all[["TPACKce"]])`, and *TPACK related to generative AI*: `r report_lmer_anova(lmers_prepost_all[["TPACKgenai"]])`.

For both subscales, we found the same pattern that students' change in Intelligent TPACK differed significantly between the control group and the two intervention groups, with students in the intervention groups having a significantly higher increase than the control group, whereas the intervention groups did not differ significantly from each other. That is, for *TPACK*:  `r report_lmer(lmers_prepost_all[["TPACKce"]], term = tc1)` and `r report_lmer(lmers_prepost_all[["TPACKce"]], term = tc2)`; and *TPACK related to generative AI*, `r report_lmer(lmers_prepost_all[["TPACKgenai"]], term = tc1)` and `r report_lmer(lmers_prepost_all[[7]], term = tc2)`.

### AI Ethics
For ethics related to the use of AI in education, we found a significant main effect of treatment on students' change in **XYZ**, `r report_lmer_anova(lmers_prepost_all[["Ethics"]])`. We found a significant higher increase in students' **XYZ** in the intervention groups compared to the control group, `r report_lmer(lmers_prepost_all[["Ethics"]], term = tc1)`, whereas the change did not differ significantly between the two intervention groups, `r report_lmer(lmers_prepost_all[["Ethics"]], term = tc2)`.

## Perceived Usefulness
Descriptive results of students' perceived usefulness of XYZ are listed in [@fig-pu], while the results of the linear mixed-effects models for perceived usefulness are presented in [@tbl-lmer-pu-summary].

```{r fig-pu, echo=FALSE, fig.cap="Descriptive Statistics of Students' Perceived Usefulness Ratings Over Time by Treatment Group.", out.width="100%",fig.width=7, fig.height=6}
p_pu +
  facet_wrap(
    ~outcome,
    ncol = 3,
    scales = "fixed",
    labeller = as_labeller(c(
      "PU" = "PU: Perceived Usefulness",
      "PU1" = "PU1: Increases Teaching Quality",
      "PU2" = "PU2: Reduces Workload",
      "PU3" = "PU3: Increases Effectiveness",
      "PU4" = "PU4: Usefulness for Work"
    ))
  ) +
  labs(y = "Rating") # + theme(legend.position = "bottom")
```

```{r  pu - contrasts}
pu_between_contrasts <- lapply(
  lmers_prepost_all[grepl("^(PU)", names(lmers_prepost_all))],
  function(m) {
    emmeans(m, ~ intervention * time) %>%
      contrast(interaction = "revpairwise", adjust = "none")
  }
)
if (!is_word) {
  print(pu_between_contrasts)
}
```

::: {.landscape}
```{r tab.layout="autofit"}
#| label: tbl-lmer-pu-summary
#| tbl-cap: "Summary of Linear Mixed-Effects Models for Perceived Usefulness."

ft_pu <- modelsummary_models(
  lmers_prepost_all[grepl("^(PU)", names(lmers_prepost_all))],
  ddf = "kr",
  output = modelsummary_output,
  # coef_rename = c("genderMännlich" = "Geschlecht")
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_pu <- ft_pu %>% adjust_table_for_word()
}
ft_pu
```

```{r pu-tab-models}
#| tbl-cap: "Summary of Linear Mixed-Effects Models for Perceived Usefulness."

lmers_pu <- lmers_prepost_all[grepl("^(PU)", names(lmers_prepost_all))]
tab_models(lmers_pu)
```

:::

For the overall perceived usefulness, we did not find a significant main effect of treatment on students' change in perceived usefulness, `r report_lmer_anova(lmers_prepost_all[["PU"]])`. Neither did the control group's change differ significantly from both intervention groups, `r report_lmer(lmers_prepost_all[["PU"]], term = tc1)`, nor did the change differ significantly between the two intervention groups, `r report_lmer(lmers_prepost_all[["PU"]], term = tc2)`.

However, when investigating the individual items of perceived usefulness, we found that students in the intervention groups showed marginally significant higher increases compared to the control group, `r report_lmer(lmers_prepost_all[["PU1"]], term = tc1)` and `r report_lmer(lmers_prepost_all[["PU2"]], term = tc1)`, whereas the change did not differ significantly between the two intervention groups, `r report_lmer(lmers_prepost_all[["PU1"]], term = tc2)` and `r report_lmer(lmers_prepost_all[["PU2"]], term = tc2)`, respectively, whereas we found no treatment effect for *PU3*  and *PU4*, see [@tbl-lmer-pu-summary].

::: {.callout-note}
wir können für uns noch die Kontraste für PU anschauen. Hier ist interessant, dass für PU1 stem > control und PU2 language > control
:::

```{r results="asis", include=F}
stargazer_models(lmers_pu)
```

# Results of ANCOVAs


We analyzed the interventions' efficacy in a multiple linear regression analysis approach. For this, we implemented a one-way ANCOVA for each outcome measure, considering our $3$ factorial design.

::: {.callout-note}
We did not include the university dummy variable for the clustering of students within universities, as we only have control students at University 2 (and thus yields a high correlation with treatment as well).
:::

Additionally, we included the respective pretest score of the outcome measures as a covariate in the analyses.
Finally, we included variables as a covariate if they met the criteria for baseline equivalence in randomized controlled trials by the What Works Clearinghouse [@WWC2022]. That is, we included the variables as covariates if the difference between the groups was statistically significant, or if it exceeded $d>.05$.
We tested for group differences concerning the following variables at baseline (pretest): age, gender, semester, and study level (Bachelor vs. Master). We found for gender and semesters  statistically significant differences between the treamtment groups, see [@tbl-participants]. We thus included the following covariates in the models: gender and semester.

That is, the final model for each outcome measure contained treatment, pretest score, gender, semester, and University as fixed effects, and was specified as follows:
`lm_robust(outcome_post ~ treatment + outcome_pre + gender + semester )`.

Posttest scores were standardized before running the analyses, so that the regression coefficients of the dummy variable indicating the respective treatment groups can directly be interpreted as Cohen’s d [@marsh2009].

The criterion of statistical significance was set at α=.05.

## Statistical Analysis
::: {.callout-note}
Die gleiche Statistik wie oben, aber statt lmer etc.:
:::

To compute heteroscedasticity-consistent (HC) standard errors, we used the *lm_robust* function from the *estimatr* package [@blair2025], specifying the HC3 estimator, as recommended for small samples [@long2000].

```{r compute-ancovas}
# reshape data to wide format for ancova
data_wide <- data %>%
  pivot_wider(
    id_cols = c(code, intervention, group, gender, semester, students, location),
    names_from = time,
    values_from = all_of(outcomes)
  ) %>% mutate(
    # scale the posttest score so that the estimates are approximations of Cohen's d
    across(all_of(outcomes %>% paste0("_post")), scale),

    treatment = factor(ifelse(intervention == "control", "control", "intervention"))
  )

contrasts.treatment <- cbind(c(-1 / 2, 1 / 2))
colnames(contrasts.treatment) <- c("-c+t")
contrasts(data_wide$treatment) <- contrasts.treatment

ancovas <- lapply(
  outcomes,
  fit_ancova,
  groups,
  # we remove University as a covariate as we only have control students at University 2
  setdiff(controls, "location"),
  "HC3",
  data_wide
)
names(ancovas) <- outcomes

ancovas_two_groups <- lapply(
  outcomes,
  fit_ancova,
  "treatment",
  setdiff(controls, "location"),
  "HC3",
  data_wide
)
names(ancovas_two_groups) <- outcomes

```

## Intelligent TPACK and AI Ethics

::: {.landscape}
```{r tab.layout="autofit"}
#| label: tbl-ancova-tpack-summary
#| tbl-cap: "Summary of Multiple Regression Models for Intelligent TPACK Subscales and AI Ethics"

ft_ancova <- modelsummary_models(
  ancovas[grepl("^(T|Ethics)", names(ancovas))],
  outcomes = outcomes %>% grep("^(T|Ethics)", ., value = TRUE),
  output = modelsummary_output,
  # coef_rename = c("genderMännlich" = "Geschlecht")
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_ancova <- adjust_table_for_word(ft_ancova, width = 1)
}
ft_ancova
```

:::

## AI Literacy

```{r tab.layout="autofit"}
#| label: tbl-ancova-ailit-summary
#| tbl-cap: "Summary of Multiple Regression Models for AI Literacy."

ft_ailit_ancova <- modelsummary_models(
  ancovas[names(ancovas) %in% c("AILIT", "AILITWolke")],
  outcomes = c("AILIT", "AILITWolke"),
  output = modelsummary_output,
  # coef_rename = c("genderMännlich" = "Geschlecht")
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_ailit_ancova <- adjust_table_for_word(ft_ailit_ancova, width = .75)
}
ft_ailit_ancova
```

```{r, results="asis"}
if (!is_word) {
  cat("### Pairwise Contrasts Between Groups")
}
```

```{r ailit-ancova-between-contrasts}
# pairwise between contrasts (no multiplicity adjustment)
ailit_ancova_between_contrasts <- lapply(ancovas[1:3], function(m) {
  emmeans(m, ~ intervention) %>%
    contrast(interaction = "pairwise", adjust = "none")
})
if (!is_word) {
  print(ailit_ancova_between_contrasts)
}
```

## Perceived Usefulness

::: {.landscape}
```{r tab.layout="autofit"}
#| label: tbl-ancova-pu-summary
#| tbl-cap: "Summary of Multiple Linear Regression Models for Perceived Usefulness."

ft_pu_ancova <- modelsummary_models(
  ancovas[grepl("^(PU)", names(ancovas))],
  outcomes = outcomes %>% grep("^(PU)", ., value = TRUE),
  output = modelsummary_output,
  # coef_rename = c("genderMännlich" = "Geschlecht")
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_pu_ancova <- ft_pu_ancova %>% adjust_table_for_word()
}
ft_pu_ancova
```

:::



## Two Groups (Control vs. Intervention)

::: {.callout-note}
Due to the very small (insignificant) difference between the distribution of gender in the treatment groups, we could drop it from the covariates.
:::


```{r effectsizes-covariates-two-groups}

# perform chi square for categorical variables
chisquare_covariates_two_groups <- vars_cat  %>% lapply(function(v) {
  tab <- table(d[[v]], d$treatment)
  chisq <- chisq.test(tab)
})
names(chisquare_covariates_two_groups) <- vars_cat

chisquare_covariates_two_groups

effectsizes_covariates__two_groups %>% kable
```

```{r ancova-two-groups}
#| label: tbl-ancova-two-groups
#| tbl-cap: "Summary of Multiple Linear Regression Models for Two Treatment Groups."

modelsummary_models(
  ancovas_two_groups,
  outcomes = outcomes,
  output = modelsummary_output
)

```

# Discussion

# Playground

## Reflection
```{r reflection}
refs <- c("Ref", list_items_ref %>% unlist)
lms_reflection <- lapply(
  refs,                        # a character vector of column names
  function(x) {
    f <- as.formula(paste(x, "~ intervention + gender + semester"))
    lm(f, data %>% filter(time == "post"))
  }
)

modelsummary_models(lms_reflection)

# pairwise between contrasts (no multiplicity adjustment)
refl_between_contrasts <- lapply(lms_reflection, function(m) {
  emmeans(m, ~ intervention) %>%
    contrast(interaction = "pairwise", adjust = "none")
})
if (!is_word) {
  print(refl_between_contrasts)
}
```
```{r plot-reflection, fig.height=8}
# for all of refs, do a boxplot of ref ~ intervention using ggplot2
data_long_refl <- data %>%
  filter(time == "post") %>%
  pivot_longer(
    cols = all_of(unname(refs)), # <-- strip names!
    names_to = "outcome",
    values_to = "score"
  )

data_long_refl <- data_long_refl %>%
  mutate(
    outcome = factor(outcome, levels = refs)
  )

# set color scale/palette and plot theme
wolke_color_scale <- scale_colour_manual(values = rev(pal_jco("default")(3)))
wolke_theme <-
  jtools::theme_apa(
    legend.pos = "bottomright",
    legend.use.title = T,
    facet.title.size = 12,
    x.font.size = 18,
    y.font.size = 18,
    legend.font.size = 12
  )

ggplot(data_long_refl, aes(x = intervention, y = score, color=intervention)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5) +
  # add mean as diamond shape
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 4,
    color = "black",
    fill = "black"
  ) +
  facet_wrap(~outcome, scales = "free_y") +
  labs(x = "Intervention Group", y = "REF") +
  wolke_color_scale +
  wolke_theme
```

## Demographics Check
```{r test}
library(dplyr)
library(tidyr)

check_demo_vars <- function(
  data,
  demo_vars = c("age", "semester", "gender", "students"),
  id = "code",
  time_col = "time",
  pre_label = "pre",
  post_label = "post",
  show_only_diff = FALSE
) {
  out <- data %>%
    select(all_of(c(id, time_col, demo_vars))) %>%
    # ensure a common type across demo vars
    mutate(across(all_of(demo_vars), as.character)) %>%
    pivot_longer(
      cols = all_of(demo_vars),
      names_to = "variable",
      values_to = "value"
      # alternatively: values_transform = as.character
      # or: values_ptypes = list(value = character())
    ) %>%
    group_by(.data[[id]], variable, .data[[time_col]]) %>%
    summarise(
      value = paste(unique(na.omit(value)), collapse = ", "),
      .groups = "drop"
    ) %>%
    pivot_wider(names_from = all_of(time_col), values_from = value) %>%
    transmute(
      !!id := .data[[id]],
      variable,
      pre = .data[[pre_label]],
      post = .data[[post_label]],
      match = (pre == post) | (is.na(pre) & is.na(post))
    ) %>%
    arrange(.data[[id]], variable)

  if (show_only_diff) {
    out <- filter(out, is.na(match) | !match)
  }
  out
}
check_demo_vars(
  data,
  demo_vars = c(
    "age",
    "semester",
    "gender",
    "students",
    "intervention",
    "location"
  ),
  id = "code",
  show_only_diff = T
)
```

### Rerun Models without Semester, but Study Level as Covariate

::: {.landscape}
```{r rerun-models-with-study-level}
lmers_prepost_all_study_level <- lapply(lmers_prepost_all, function(m) {
  update(m, . ~ . - semester + students)
})
modelsummary_models(
  lmers_prepost_all_study_level,
  ddf = "kr",
  output = modelsummary_output
)
```

::: 

## Comparison of Certain Covariates

### In- and Exclusion of University as Covariate
```{r}

modelsummary_models(list(ancovas[["AILITWolke"]], 
lm_robust(AILITWolke_post ~ intervention + pretest + gender + semester + location, data_wide %>% mutate(pretest = AILITWolke_pre), se_type="HC3"),
lm(AILITWolke_post ~ intervention + pretest + gender + semester, data_wide %>% mutate(pretest = AILITWolke_pre)),
lm_robust(AILITWolke_post ~ treatment + pretest + gender + semester , data_wide %>% mutate(pretest = AILITWolke_pre), se_type="HC3")),
  output = modelsummary_output)
```

## LMERs with Standardized Outcomes
```{r lmers-standardized}
modelsummary_models(list(lmers_prepost_all[["AILITWolke"]],
lmer(AILITWolke~ intervention * time +(1|code), data),
lmer(AILITWolke~ intervention * time +(1|code), data %>% group_by(time) %>% mutate(AILITWolke = scale(AILITWolke)) %>% ungroup()))
, output = modelsummary_output
)
```

## Cluster-Robust Standard Errors for LMER Models

::: {.callout-tip}
If we used cluster-robust standard errors in the lmer models, the SEs become a little smaller, resulting in slightly smaller p-values.
:::

```{r tbl-lmer-robust-se}
#| tbl-cap: "Summary of Linear Mixed-Effects Models with Cluster-Robust Standard Errors and Satterthwait-Corrected Tests Using clubSandwich::vcovCR with CR2."


# Cluster-robust vcov (CR2 is usually recommended)
# However, modelsummary itself doesn't directly implement Satterthwaite tests; it relies on the vcov matrix for standard errors and calculates tests using default methods. 

modelsummary_models(lmers_prepost_all, vcov = "CR2")

# Alternative, which is not needed if modelsummary correctly derives the cluster variable from the model (which is the case for the random intercept of code)
vcov_list <- outcomes %>% lapply(function(o) {
  m <- lmers_prepost_all[[o]]
  # cluster-robust variance estimator with CR2 small-sample correction
  clubSandwich::vcovCR(m, cluster = data$code, type = "CR2")
})
# modelsummary_models(lmers_prepost_all, vcov = vcov_list)
```


```{r tbl-lmer-robust-se-satt}
#| tbl-cap: "Summary of Linear Mixed-Effects Models with Cluster-Robust Standard Errors and Satterthwaite-Corrected Tests Using coef_tests."


# Extract coefficient tests with cluster-robust CR2 SE and Satterthwaite correction
coef_tests <- lapply(lmers_prepost_all, function(model) {
  clubSandwich::coef_test(model, vcov = "CR2", cluster = data$code, test = "Satterthwaite")
})
# Simple custom class for robust results
create_robust_lmer <- function(model, coef_test_result) {
  robust_obj <- structure(
    list(
      model = model,
      robust_results = coef_test_result
    ),
    class = c("robust_lmer", class(model))
  )
  
  # Copy important attributes from the original model
  attr(robust_obj, "formula") <- formula(model)
  
  return(robust_obj)
}

# Simple tidy method - just return your robust results
tidy.robust_lmer <- function(x, ...) {
  ct <- x$robust_results
  data.frame(
    term = rownames(ct),
    estimate = ct$beta,
    std.error = ct$SE,
    statistic = ct$tstat,
    p.value = ct$p_Satt,
    stringsAsFactors = FALSE
  )
}

# Simple glance method - get basic stats + R²
glance.robust_lmer <- function(x, ...) {
  model <- x$model
  
  # Basic model stats
  basic_stats <- data.frame(
    nobs = nobs(model),
    aic = AIC(model),
    bic = BIC(model),
    logLik = as.numeric(logLik(model)),
    Std.Errors = "CR2"
  )
  
  # Add R² marginal and conditional
    # Add R² using custom names that will display as you want
  if (requireNamespace("performance", quietly = TRUE)) {
    r2 <- performance::r2(model)
    basic_stats$`R2 Marg.` <- r2$R2_marginal %>% round(3)       
    basic_stats$`R2 Cond.` <- r2$R2_conditional %>% round(3)
  }
  return(basic_stats)
}

# Create robust models
robust_models <- lapply(seq_along(lmers_prepost_all), function(i) {
  create_robust_lmer(lmers_prepost_all[[i]], coef_tests[[i]])
})

# Use with modelsummary
modelsummary_models(
  robust_models
)
```

