---
title: "WoLKE-Pilot-2025"
author: "Benedikt Beuttler"
date: "`r Sys.Date()`"
lang: de
page-layout: full
document-css: false
link-citations: true
number-sections: true
editor: visual
execute:
  cache: true
format:
  html:
    standalone: true
    embed-resources: true
    section-divs: true
    html-math-method: mathjax
    wrap: none
    default-image-extension: png
    date-format: long
---

# Preamble

## Prepare the Environment

## Load Libraries

```{r load-libraries-and-functions}
#| output: false
library(this.path)
library(dplyr) #e.g. for select
library(tidyr) # for reshaping data
library(stringdist) #levenshtein
library(stringr) # for reshaping data
library(lme4)
library(lmerTest)
library(emmeans)
library(modelsummary)
library(kableExtra)
library(arsenal)
library(forcats)
source(file = "functions.r") # all outsourced funtions
```

## Load Pre- and Posttest and Reshape Data

```{r load-data}
# Set working directory to the script's location
setwd(this.path::this.dir())
### Load Data
# load csv -> df_pre
source("./data/import_ki-lehramt.r")

# extract items starting with TP (TPACK) or DE (Demographics)
data_raw <- df %>%
  select(
    QUESTNNR,
    starts_with("DE"),
    starts_with("AI"),
    starts_with("ED"),
    starts_with("TP"),
    starts_with("RE"),
    starts_with("CT"),
    starts_with("STARTED"),
    starts_with("FINISHED")
  )
# sum(data_raw$QUESTNNR == "mint-post") # 10 are incorrectly labeled
data_raw$QUESTNNR[
  substr(data_raw$STARTED, 6, 7) == "07" &
    grepl("mint", data_raw$QUESTNNR)
] <- "mint-post" # fixe some wrong entries in pre list to post

# rename demographics
data_raw <- data_raw %>%
  mutate(
    code = tolower(DE13_01),
    age = DE14_01,
    semester = DE15_01,
    studies = DE16,
    subject1 = DE24,
    subject2 = DE25,
    gender = DE10,
    location = DE11,
    semestercount = DE15_01,
    isp = DE09,
    #sprache_intervention = DE08,
    sprache_last_sem = DE06, #already visited cans KI seminar
    #mint_control = DE07,
    mint_exclusion = DE02,
    intervention_sprache = case_when(
      str_to_lower(as.character(DE08)) == "ja" ~ TRUE,
      str_to_lower(as.character(DE08)) == "nein" ~ FALSE
    ),
    intervention_mint = case_when(
      str_to_lower(as.character(DE07)) == "ja" ~ TRUE,
      str_to_lower(as.character(DE07)) == "nein" ~ FALSE
    ),
    intervention_only = coalesce(intervention_sprache, FALSE) |
      coalesce(intervention_mint, FALSE),
    intervention_only = as.logical(intervention_only)
  )
data_raw <- data_raw[, c("code", setdiff(names(data_raw), "code"))] #set code as first column
# View(data_raw[(is.na(data_raw$code) | data_raw$code == ""), ])
data_raw <- data_raw[!(is.na(data_raw$code) | data_raw$code == ""), ] # remove empty/na codes (=27)

data_raw_clean <- data_raw

# factorize some columns
data_raw_clean$gender <- factor(data_raw_clean$gender) # TODO: umbenennen und reihenfolge der levels angeben
data_raw_clean$location <- factor(data_raw_clean$location) # TODO: umbenennen und reihenfolge der levels angeben

# merge pre und post dfs
data <- data_raw_clean

# create relevant columns group (sprache mint), time (pre, post) and intervention (control, mint, sprache)
data <- data %>%
  mutate(
    group = case_when(
      str_detect(QUESTNNR, "sprache") ~ "sprache",
      str_detect(QUESTNNR, "mint") ~ "mint"
    ),
    time = case_when(
      str_detect(QUESTNNR, "post") ~ "post",
      TRUE ~ "pre"
    ),
    group = factor(group), # TODO: mint=stem sprache=language
    time = factor(time, levels = c("pre", "post"))
  )

# create intervention with 3 factors: control, mint, sprache
data <- data %>%
  mutate(
    intervention = factor(ifelse(
      intervention_only == F,
      "control",
      ifelse(group == "mint", "mint", "sprache") # TODO: mint=stem sprache=language
    ))
  )

### EXCLUSIONS
e1 <- nrow(data[data$FINISHED == F, ]) # not FINISHED
e1
data <- data[data$FINISHED == T, ] # only keep completed questionnaires

e2 <- nrow(data[data$code == "höel07", ])
e2
data <- data[data$code != "höel07", ] #example code

# check for multiple entries and completed questionnaires
# table(table(data$code))
# view a specific code; obsu02 stfr08 (both multiple finished questionnaire, rowindex 1811)
data <- data[!rownames(data) %in% c("1811", "1814"), ] # Wie haben wir uns entschieden? am besten erster eintrag nehmen
# View(data[data$code == "obsu02", ])

# Code corrections from levenshtein checks
data$code[data$code == "mair10"] <- "moir10"
data$code[data$code == "drch17"] <- "frch17"
data$code[data$code == "loma26"] <- "goma26"
data$code[data$code == "sisu00"] <- "sisu01"
data$code[data$code == "mosy24"] <- "stsy24"
data$code[data$code == "frco30"] <- "scco30"
data$code[data$code == "meli06"] <- "meli63"
data$code[data$code == "waol30"] <- "spol30"

# check code distances
check_dist <- F
if (check_dist) {
  check_distances(data, 2)
}

data_all <- data
# save data
save(data_all, file = "data/data_all.Rdata")

# only keep subjects with both data in pre and post test for analysis
data <- data %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  ungroup()

# investigate: df with subjects having unclear studies, check semestercount to see if BA or MA
temp_df <- data %>%
  filter(!is.na(DE16_16)) %>%
  select(code, DE16_16, semester, intervention)

# append (Master) to those with semestercount > 6 according to tmp_df
data$DE16_16[data$code %in% temp_df$code] <-
  paste0(data$DE16_16[data$code %in% temp_df$code], " (Master)")

remove(temp_df)

# create new variable subject with levels bachelor and master
data <- data %>%
  mutate(
    students = case_when(
      str_detect(tolower(studies), "master") ~ "master",
      str_detect(tolower(studies), "bachelor") ~ "bachelor",
      str_detect(tolower(studies), "nachstudium") ~ "master",
      str_detect(tolower(studies), "sonstiges") &
        str_detect(tolower(DE16_16), "master") ~
        "master",
      str_detect(tolower(studies), "sonstiges") &
        str_detect(tolower(DE16_16), "bachelor") ~
        "bachelor",
      # TRUE ~ "master" # hier gibt es eine person, die nachstudium angegeben hat, aber mit semestercount 6/8 setzen wir ihn auf master
      TRUE ~ NA
    ),
    students = factor(students, levels = c("bachelor", "master"))
  )

save(data, file = "data/data.Rdata")

```

```{r Demographics}
# Generate the tableby summary
preprocess_data <- function(data_all) {
  data %>%
    droplevels() %>%
    mutate(
      studies = fct_infreq(studies),
      subject1 = fct_infreq(subject1),
      subject2 = fct_infreq(subject2)
    )
}

# only get data from pretest for demographic data to see registered
data_pretest <- data_all %>%
  filter(time == "pre")
nrow(data_pretest) # amount of registered participants 145

# only take samples with pre and post test data into account, singlecodes
data_dem_all <- data_all %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  filter(time == "post") %>%
  ungroup()
nrow(data_dem_all) # 76 participants with pre and post test

# Create the formula
demographics_formula <- intervention ~
  location + age + gender + semester + isp + studies + subject1 + subject2
demographics <- tableby(demographics_formula, data = data_dem_all)

# Ausgabe als Tabellen
summary(demographics) %>%
  as.data.frame() %>%
  knitr::kable(caption = "Demographic Summary")

# Welche Fächer gibt es wie oft?
counts <- data_dem_all %>%
  pivot_longer(
    c(subject1, subject2),
    names_to = "slot",
    values_to = "subject"
  ) %>%
  mutate(subject = str_squish(subject)) %>%
  filter(!is.na(subject), subject != "") %>%
  count(subject, sort = TRUE, name = "n") %>%
  mutate(p = n / sum(n))

counts

```

```{r Prepare scales}

### grouping scales
## TPACK
# all TPACK items used
list_items_i_tpack <- list(
  "TK" = grep("^TP01_", names(data), value = T),
  "TPK" = grep("^TP02_", names(data), value = T),
  "TCK" = grep("^TP03_", names(data), value = T),
  "TPACK_know" = grep("^TP04_(04|05|06|07|08|15)$", names(data), value = T), # Sub-Subskala 1: Wissen ("ich weiß, wie ...")
  "TPACK_comp" = grep(
    "^TP04_(01|02|03|09|10|11|12|13|14)$",
    names(data),
    value = T
  ), # Sub-Subskala 2: Kompetenzen ("ich kann ...")
  "TPACK_genai" = grep("^TP04_(09|10|11|13|14)$", names(data), value = T), # Sub-Subskala 3: GenAI Items
  "Ethics" = grep("^TP05_0[1-6]$", names(data), value = T) #im pretest nur items 1-6 vorhanden 7,8 kamen im post dazu
)

# ein paar ethik items müssen noch rausgenommen werden, die nur im posttest erhoben wurden. hier nur check
# data_subset1 <- data %>% select(all_of(unlist(list_items_i_tpack)))
# View(data_subset1)

## Perceived Usefulnes
list_items_pu <- list(
  "Pu" = grep("^ED03_", names(data), value = T)
)
# data_subset2 <- data %>% select(all_of(unlist(list_items_pu)))
# sum(data_subset2 < 0, na.rm = TRUE)

## Reflection (bei Piloten irrelevant)
# list_items_ref <- list(
#   "Ref" = grep("^RE01_", names(data), value=T)
# )
# data_subset3 <- data %>% select(all_of(unlist(list_items_ref)))
# sum(data_subset3 < 0, na.rm = TRUE)

## AI Literacy
# data <- data %>% mutate(AILIT = AI35_02) # rename to sumscore
list_items_ailit_wolke <- list(
  "ailit_wolke" = grep(
    "^(AI02|AI05|AI06|AI10|AI13|AI14|AI15|AI17_0[1-5]|AI18|AI20|AI21|AI23|AI24|AI25|AI27|AI28|AI29)",
    names(data),
    value = T
  ), # items we think students learn in the WoLKE seminar
  "ailit_short" = grep(
    "^(AI01|AI04|AI09|AI015|AI17_0[1-5]|AI19|AI21|AI25|AI27|AI29)$",
    names(data),
    value = T
  ) # short version of the AILIT
)

# create new columns with the means of the lists
data <- data %>%
  mutate(
    AILIT = AI35_02, # default sumscore
    AILITWoLKE = rowMeans(
      select(., all_of(list_items_ailit_wolke$ailit_wolke)),
      na.rm = TRUE
    ),
    AILITShort = rowMeans(
      select(., all_of(list_items_ailit_wolke$ailit_short)),
      na.rm = TRUE
    ),
    TK = rowMeans(select(., all_of(list_items_i_tpack$TK)), na.rm = TRUE),
    TPK = rowMeans(select(., all_of(list_items_i_tpack$TPK)), na.rm = TRUE),
    TCK = rowMeans(select(., all_of(list_items_i_tpack$TCK)), na.rm = TRUE),
    TPACKknow = rowMeans(
      select(., all_of(list_items_i_tpack$TPACK_know)),
      na.rm = TRUE
    ),
    TPACKcomp = rowMeans(
      select(., all_of(list_items_i_tpack$TPACK_comp)),
      na.rm = TRUE
    ),
    TPACKgenai = rowMeans(
      select(., all_of(list_items_i_tpack$TPACK_genai)),
      na.rm = TRUE
    ),
    Ethics = rowMeans(
      select(., all_of(list_items_i_tpack$Ethics)),
      na.rm = TRUE
    ),
    PU = rowMeans(select(., all_of(list_items_pu$Pu)), na.rm = TRUE),
    # Ref    = rowMeans(select(., all_of(list_items_ref$Ref)), na.rm = TRUE),
    TPACK_total = rowMeans(
      select(., all_of(unlist(list_items_i_tpack))),
      na.rm = TRUE
    )
  )

```

## Analysis

Im folgenden verwenden wir drei Kontraste: (0) Intercept = Gesamtdurchschnitt aller Gruppen, (1) Exp-Ø (MINT+Sprache)/2 vs. Control, (2) Sprache vs. MINT. Mit der Zeit-Interaktion testen wir, ob sich diese Unterschiede über die Zeit unterschiedlich entwickeln (Difference-in-Differences: Exp-Ø vs. Control sowie Sprache vs. MINT).
``` {r lmer MINT vs Sprache vs Kontrollgruppe}
## define contrast Control vs Exp1+Exp2 - vermutlich bessere version, da die Estimates "stimmen"
mat <- cbind(
  c(1 / 3, 1 / 3, 1 / 3), # Durchschnitt aller 3 Gruppen
  c(-1, .5, .5), # Vergleich Control vs Mittel der beiden Interventions-Gruppen
  c(0, -1, 1)
) # Vergleich Sprache vs MINT
contrasts.CvsExp <- solve(t(mat))[, -1]
colnames(contrasts.CvsExp) <- c("-C+MINT_Sprache", "-MINT+Sprache")
contrasts(data_all$intervention) <- contrasts.CvsExp

# or use this coding system?
# my.helmert <- matrix(c(-2 / 3, 1 / 3, 1 / 3, 0, -1 / 3, 1 / 3), ncol = 2)
# colnames(my.helmert) <- c("-C+MINT_Sprache", "-MINT+Sprache")
# contrasts(data$cond2) <- my.helmert

## define variables
random_terms_lmer <- c("(1 | code)")
groups <- c("intervention")
outcomes <- c(
  "AILIT",
  "TK",
  "TPK",
  "TCK",
  "TPACKknow",
  "TPACKcomp",
  "TPACKgenai",
  "Ethics",
  "PU"
)
time <- "time"
controls <- c("age", "gender", "location")

# groups set as NA
lmers_prepost_all <-
  lapply(
    outcomes,
    fit_lmer,
    groups,
    time,
    controls,
    random_terms_lmer,
    NULL,
    data = data
  )

modelsummary_models(lmers_prepost_all)


# ---------- klappt nich lol
# data_sg <- data %>%
#   group_by(code) %>%
#   filter(all(c("pre", "post") %in% time)) %>%
#   filter(location=="PH Schwäbisch Gmünd") %>%
#   ungroup()
# controls <- c("age", "gender")

# lmers_prepost_sg <-
#   lapply(outcomes, lmer, NULL, time, controls, random_terms_lmer, NULL, data = data_sg)

# modelsummary_models(lmers_prepost_sg)

# mm <- lmer(PU ~ time*intervention + age + gender + (1 | code), data = data %>% filter(all(c("pre", "post") %in% time)))
# mm %>% summary(corr=F)
# emmeans(lmers_prepost_all[[1]], ~ intervention*time|location, lmer.df = "satterthwaite") %>% contrast(interaction = "pairwise", adjust="none")
```

```{r lmer-Sprache}

#### SPRACHE ####
# nur daten berücksichtigen die pre und post test haben
data_sprache <- data %>%
  filter(group == "sprache") %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  ungroup()

## define variables
random_terms_lmer <- c("(1 | code)")
# groups <- c("intervention")
outcomes <- c(
  "AILIT",
  "TK",
  "TPK",
  "TCK",
  "TPACKappl",
  "TPACKintrg",
  "TPACKgenai",
  "Ethics",
  "PU"
)
time <- "time"
controls <- c("age", "gender")

# groups set as NA
lmers_prepost <-
  lapply(
    outcomes,
    fit_lmer,
    NULL,
    time,
    controls,
    random_terms_lmer,
    NULL,
    data = data_sprache
  )

modelsummary_models(lmers_prepost)
```
