---
title: "WoLKE-Studie-2026"
author: "Benedikt Beuttler"
date: "`r Sys.Date()`"
lang: en
page-layout: full
document-css: false
link-citations: true
number-sections: true
toc: true
csl: https://www.zotero.org/styles/apa
bibliography: library.bib
editor: source
execute:
  cache: true
format:
  html:
    standalone: true
    embed-resources: true
    section-divs: true
    html-math-method: mathjax
    wrap: none
    default-image-extension: png
    date-format: long  
    code-fold: true
    code-tools: true
  docx: 
    toc: true
    toc-depth: 2
    number-sections: true
    fig-caption: true
    fig-format: pdf
    default-image-extension: pdf
    date-format: long
    keep-md: true
    echo: false
    code-fold: true
prefer-html: true
---


This document contains the analysis scripts and report for the manuscript "A Course on AI for Pre-Service STEM and Language Teachers: A Quasi-Experimental Evaluation of Learning Outcomes Using Intelligent-TPACK and AI Literacy Instruments ". The R scripts, data are hosted at <OSF>.

```{r Information for first time user to get started with renv}
#| output: false
if (FALSE) {
  install.packages("renv") #1
  renv::restore() #2
}
```

# Preamble

## Prepare the Environment

## Load Libraries

```{r load-libraries-and-functions}
#| output: false

# install pacman
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}
library(pacman)

## Pfad/Utilities
p_load(this.path)

## Wrangling
# use pacman to load packages
p_load(dplyr, tidyr, stringr, forcats, purrr, magrittr)

## Ähnlichkeit/Text
p_load(stringdist, psych)

## Modellierung
p_load(lavaan, lme4, lmerTest, emmeans, estimatr)

## Tabellen/Reporting
p_load(modelsummary, kableExtra, arsenal)

## Visualisierung/Themes
p_load(ggplot2, ggsci, jtools)

p_load(mnormt)

source(file = "functions.r") # all outsourced funtions
```

```{r prepare-output}

is_word <- !(knitr::is_html_output() | knitr::is_latex_output())

if (is_word) {
  p_load(flextable)
}

```


## Load Pre- and Posttest and Reshape Data

```{r load and prepare data 1}
# if there is env-analysis.Rdata, load that to get the data frame df, otherwise run this cell
if (FALSE) {
  # file.exists("env-analysis.Rdata")) {
  load("data/env-analysis.Rdata")
} else {
  # Set working directory to the script's location
  # safe working-directory selection (replace setwd(this.path::this.dir()))
  wd <- tryCatch(this.path::this.dir(), error = function(e) NULL)

  if (is.null(wd)) {
    wd <- tryCatch(
      {
        if (
          requireNamespace("rstudioapi", quietly = TRUE) &&
            rstudioapi::isAvailable()
        ) {
          dirname(rstudioapi::getActiveDocumentContext()$path)
        } else if (requireNamespace("here", quietly = TRUE)) {
          here::here()
        } else {
          getwd()
        }
      },
      error = function(e) getwd()
    )
  }
  setwd(normalizePath(wd))

  ### Load Data
  # load csv -> df_pre
  source("./data/import_data.r")

  # extract relevant items in a separate data frame
  data_raw <- df %>%
    select(
      QUESTNNR,
      starts_with("DE"), # Demographics
      starts_with("AI"), # AI Literacy Hornberger
      starts_with("GL"), # AI Literacy GLAT
      starts_with("TP"), # TPACK
      starts_with("RE"), # Reflexion
      starts_with("CT"), # Computational Thinking
      starts_with("STARTED"),
      starts_with("FINISHED")
    )
  # replace _posttest with _post for easier handling
  data_raw$QUESTNNR <- str_replace(
    data_raw$QUESTNNR,
    "_posttest",
    "_post"
  )
  # rename demographics
  data_raw <- data_raw %>%
    mutate(
      code = tolower(DE13_01),
      age = DE14_01,
      semester = DE15_01,
      studies = DE16,
      subject1 = DE24,
      subject2 = DE25,
      gender = DE10,
      location = DE11,
      semestercount = DE15_01,
      isp = DE09,
      ai_training = DE32,
      ai_training_amount = DE33,
      #sprache_intervention = DE08,
      sprache_last_sem = DE06, #already visited cans KI seminar
      #mint_control = DE07,
      mint_exclusion = DE02,
      intervention_sprache = case_when(
        str_to_lower(as.character(DE08)) == "ja" ~ TRUE,
        str_to_lower(as.character(DE08)) == "nein" ~ FALSE
      ),
      intervention_mint = case_when(
        str_to_lower(as.character(DE07)) == "ja" ~ TRUE,
        str_to_lower(as.character(DE07)) == "nein" ~ FALSE
      ),
      intervention_only = coalesce(intervention_sprache, FALSE) | # is intervention group (stem or language)
        coalesce(intervention_mint, FALSE),
      intervention_only = as.logical(intervention_only)
    )
  data_raw <- data_raw[, c("code", setdiff(names(data_raw), "code"))] #set code as first column

  # create relevant columns group (sprache mint), time (pre, post) and intervention (control, mint, sprache)
  data_raw <- data_raw %>%
    mutate(
      group = case_when(
        str_detect(QUESTNNR, "sprache") ~ "language",
        str_detect(QUESTNNR, "mint") ~ "stem"
      ),
      time = case_when(
        str_detect(QUESTNNR, "post") ~ "post",
        TRUE ~ "pre"
      ),
      group = factor(group, levels = c("language", "stem")),
      time = factor(time, levels = c("pre", "post"))
    )

  # create intervention with 3 factors: control, mint, sprache
  data_raw <- data_raw %>%
    mutate(
      intervention = factor(
        ifelse(
          intervention_only == F,
          "control",
          ifelse(group == "stem", "stem", "language")
        ),
        levels = c("control", "stem", "language")
      )
    )
  # create new variable subject with levels bachelor and master depending on studies value
  data_raw <- data_raw %>%
    mutate(
      students = case_when(
        str_detect(tolower(studies), "master") ~ "master",
        str_detect(tolower(studies), "bachelor") ~ "bachelor",
        TRUE ~ NA
      ),
      students = factor(students, levels = c("bachelor", "master"))
    )
}
  ``` 

``` {r checks, corrections and exclusions}

## pre-correction exclusions
# exclude, count and save the count of the ones that have not finished the questionnaire (FINISHED == F) and check how many that are
total_n_before_exclusion <- nrow(data_raw) # 345 total
not_finished_count <- sum(data_raw$FINISHED == "FALSE") # 46 removed
data_raw <- data_raw %>% filter(FINISHED == "TRUE") # 299 remaining
# count and exclude the one that have TRUE in mint_exclusion
mint_exclusion_count <- sum(data_raw$mint_exclusion == "Ja", na.rm = TRUE) # 28 removed
data_raw <- data_raw %>% filter(is.na(mint_exclusion) | mint_exclusion != "Ja") # 271 remaining

## checks
# check demograph. differences pre to post
dem_diff <- check_demo_vars(
  data_raw,
  demo_vars = c(
    "age",
    "semester",
    "gender",
    "students",
    "studies",
    "intervention",
    "location"
  ),
  id = "code",
  show_only_diff = T
)
## corrections
# age
codes_age <- dem_diff %>%
  filter(variable == "age") %>%
  pull(code)
# correct age in post to age in pre (to report one age and have same age for subj when calc models)
for (cc in codes_age) {
  # save pre value
  pre_val <- data_raw$age[data_raw$code == cc & data_raw$time == "pre"]

  # write that in post value
  data_raw$age[data_raw$code == cc & data_raw$time == "post"] <- pre_val
}

# TODO Sina?
# collect codes which need to be individually checked, and what for
codes_to_check <- c(
  "enan27", # age, semester, studies 2 mal?
  "enta20", # age, 2 mal?
  "enan26", # bachelor/master
  "ennu03", # semester
  "enro23", # semester
  "ennu31", # semester
  "enro23", # studies
  "enut12", # studies
  "enut20", # verschreiber mit enut12?
  "imel28", # studies
  "llsa08", # studies
  "llsa09", # studies
  "llsi08", # studies
  "msma08", # semester
  "rgal31", # semester
  "rgma26" # semester
)

# view all codes_to_check in the raw data to check in a separate df (just take data_raw for that)
View(data_raw %>% filter(code %in% codes_to_check))
View(data_raw %>% filter(code == "enta20"))


## exclusion
# count in separate variable and exclude those with NA in code or empty code
na_in_code <- sum(is.na(data_raw$code) | data_raw$code == "") # 33 removed
data_raw <- data_raw[!(is.na(data_raw$code) | data_raw$code == ""), ] # 266 remaining


## checks
# check data types of all non test related columns
str(data_raw[, c(1, 2, 184:204)])

# data_raw$STARTED has this format "2026-01-27 12:21:13 CET" but i wanna check with tables only the dates, not the time
date_check <- as.Date(substr(data_raw$STARTED, 1, 10))
table(date_check)


data <- clean_df(data_raw)
# data_raw_clean <- data_raw

## factorize some columns
# define new levels for gender:
# levels(data_raw_clean$gender) <- c("diverse", "no answer", "male", "female")
# change order of levels
data_raw_clean$gender <- factor(
  data_raw_clean$gender
  # levels = c("female", "male", "diverse", "no answer")
)

# for location
data_raw_clean$location <- factor(
  data_raw_clean$location,
  levels = c("PH Ludwigsburg", "PH Schwäbisch Gmünd")
)

#count how many negative values there are in all RE and TP columns
neg_counts <- colSums(
  dplyr::select(data_raw_clean, dplyr::starts_with(c("RE", "TP"))) < 0,
  na.rm = TRUE
)

# # which should not be negative and replace them with NA
# data_raw_clean <- data_raw_clean %>%
#   mutate(across(
#     starts_with(c("RE", "TP")),
#     ~ ifelse(. < 0, NA, .)
#   ))

# # identify columns of interest
# cols_to_check <- grep(
#   "^(CT|TP|AI36|ED|RE)",
#   names(data_raw_clean),
#   value = TRUE
# )

# # reshape to long format and check for negatives
# negatives_df <- data_raw_clean %>%
#   select(code, all_of(cols_to_check)) %>% # keep ID/code and relevant columns
#   pivot_longer(
#     cols = all_of(cols_to_check),
#     names_to = "variable",
#     values_to = "value"
#   ) %>%
#   filter(value < 0)

# # list of codes where a negative was found
# codes_with_negatives <- unique(negatives_df$code)

# # count how many negatives per code
# counts_per_code <- negatives_df %>%
#   count(code, name = "n_negatives")

# # total number of negative instances
# total_negatives <- nrow(negatives_df)

# outputs:
# codes_with_negatives # vector of codes
# counts_per_code # table of how many negatives per code
# total_negatives # overall number of negative values

# # replace negatives with NA in the original dataframe
# data_raw_clean <- data_raw_clean %>%
#   mutate(across(all_of(cols_to_check), ~ ifelse(. < 0, NA, .)))

# merge pre und post dfs
data <- data_raw_clean

# following codes to be excluded (from Ute) due to various reasons (e.g., not taking part in seminar, moved, or already visited KI seminar before)
codes_to_exclude <- c(
  "chgi15",
  "enma23",
  "rtev15",
  "enje10"
)
# data <- data[data$code != "höel07", ] #example code

# check for multiple entries and completed questionnaires
table(table(data$code))
# find out which codes have multiple entries
codes_multiple <- names(which(table(data$code) > 2)) # "enan27" "enta20"
# view a specific code; obsu02 stfr08 (both multiple finished questionnaire, rowindex 1811)
# rownames(data)[data$code == "enan27"]
# for now exclude those two from the data "enan27" "enta20", extend codes_to_exclude with those two
codes_to_exclude <- c(codes_to_exclude, c("enan27", "enta20"))
data <- data[!data$code %in% codes_to_exclude, ]

# data <- data[!rownames(data) %in% c("1811", "1814"), ] # TODO: Wie haben wir uns entschieden? am besten erster eintrag nehmen
# View(data_raw[data_raw$code == "obsu02", ]) #stfr08

## Corrections

# Code corrections from levenshtein checks
# data$code[data$code == "mair10"] <- "moir10"
# data$code[data$code == "drch17"] <- "frch17"
# data$code[data$code == "loma26"] <- "goma26"
# data$code[data$code == "sisu00"] <- "sisu01"
# data$code[data$code == "mosy24"] <- "stsy24"
# data$code[data$code == "frco30"] <- "scco30"
# data$code[data$code == "meli06"] <- "meli63"
# data$code[data$code == "waol30"] <- "spol30"

# investigate: df with subjects having unclear studies, check semestercount to see if BA or MA
# temp_df <- data %>%
#   filter(!is.na(DE16_16)) %>%
#   select(code, DE16_16, semester, intervention)

# # append (Master) to those with semestercount > 6 according to tmp_df
# data$DE16_16[data$code %in% temp_df$code] <-
#   paste0(data$DE16_16[data$code %in% temp_df$code], " (Master)")

# remove(temp_df)

# check code distances
# check_dist <- F
# if (check_dist) {
#   check_distances(data, 2)
# }

# check demograph. differences pre to post
dem_diff <- check_demo_vars(
  data,
  demo_vars = c(
    "age",
    "semester",
    "gender",
    "students",
    "intervention",
    "location"
  ),
  id = "code",
  show_only_diff = T
)

# correct students in post to students in pre (to report one value and have same student status for subj when calc models)
codes_students <- dem_diff %>%
  filter(variable == "students") %>%
  pull(code)
for (cc in codes_students) {
  # save pre value
  pre_val <- data$students[data$code == cc & data$time == "pre"]

  # write that in post value
  data$students[data$code == cc & data$time == "post"] <- pre_val
}

codes_gender <- dem_diff %>%
  filter(variable == "gender") %>%
  pull(code)
# correct gender in post to gender in pre
for (cc in codes_gender) {
  # save pre value
  pre_val <- data$gender[data$code == cc & data$time == "pre"]

  # write that in post value
  data$gender[data$code == cc & data$time == "post"] <- pre_val
}

data$gender <- factor(
  data$gender
)

# # look at data entries for codes kesü16, leel08, mail19, obpe30, osel08, rosa20, röma24, scna18, stfr08
# codes_of_interest <- c(
#   "rosa20"
# )
# df_subset <- data %>%
#   filter(code %in% codes_of_interest) %>%
#   select(
#     code,
#     subject1,
#     subject2,
#     students,
#     studies,
#     gender,
#     age,
#     location,
#     semester,
#     time,
#     intervention,
#     isp
#   )
# View(df_subset)

# semester count correction (post to pre) after reviewing dem. data
adjust_semester_to_pre <- c(
  "enbi16",
  "enca28",
  "enhe07",
  "enke20",
  "ennu03",
  "enze11",
  "ensa03",
  "inca02",
  "llsa08",
  "llsi08",
  "llsa09",
  "msma08",
  "nnbe18",
  "rgma26",
  "ttqu16",
  "ennu31",
  "enro23",
  "lmul12",
  "rgal31"
)
for (cc in adjust_semester_to_pre) {
  # save pre value
  pre_val <- data$semester[data$code == cc & data$time == "pre"]

  # write that in post value
  data$semester[data$code == cc & data$time == "post"] <- pre_val
}

# semester count correction (pre to post) after reviewing dem. data
# adjust_semester_to_post <- c("kesü16", "stfr08", "obpe30", "mail19")
# for (cc in adjust_semester_to_post) {
#   # save pre value
#   pre_val <- data$semester[data$code == cc & data$time == "post"]

#   # write that in post value
#   data$semester[data$code == cc & data$time == "pre"] <- pre_val
# }

data_all <- data
# save data for demographics (no pre-post filtering)
save(data_all, file = "data/data_all.Rdata")

# only keep subjects with both data in pre and post test for analysis
data <- data %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  ungroup()

save(data, file = "data/data.Rdata")

```

```{r Demographics}
# Generate the tableby summary
preprocess_data <- function(data_all) {
  data %>%
    droplevels() %>%
    mutate(
      studies = fct_infreq(studies),
      subject1 = fct_infreq(subject1),
      subject2 = fct_infreq(subject2)
    )
}

# only get data from pretest for demographic data to see registered
# data_pretest <- data_all %>%
#   filter(time == "pre")

# nrow(data_pretest) # amount of registered participants 145

# only take samples with pre and post test data into account, singlecodes
data_dem_all <- data_all %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  filter(time == "post") %>%
  ungroup()
nrow(data_dem_all) # 78 participants with pre and post test

# Create the formula
demographics_formula <- intervention ~
  location + age + gender + semester + isp + studies #+ subject1 + subject2
demographics <- tableby(demographics_formula, data = data_dem_all)

# Ausgabe als Tabellen
summary(demographics) %>%
  as.data.frame() %>%
  knitr::kable(caption = "Demographic Summary")

# Welche Fächer gibt es wie oft?
subject_counts <- data_dem_all %>%
  group_by(intervention) %>%
  pivot_longer(
    c(subject1, subject2),
    names_to = "slot",
    values_to = "subject"
  ) %>%
  mutate(subject = str_squish(subject)) %>%
  filter(!is.na(subject), subject != "") %>%
  count(subject, sort = TRUE, name = "n") #%>%
# mutate(p = n / sum(n))

print(n = 100, subject_counts)

```

```{r Prepare scales}

### grouping scales
## TPACK
# all TPACK items used
list_items_i_tpack <- list(
  "TK" = grep("^TP01_", names(data), value = T),
  "TPK" = grep("^TP02_", names(data), value = T),
  "TCK" = grep("^TP03_", names(data), value = T),
  "TPACK_ce" = grep("^TP04_(0[1-8]|12|15)$", names(data), value = T), # Sub-Subskala 1: Celik Enhanced Items
  "TPACK_genai" = grep("^TP04_(09|10|11|13|14)$", names(data), value = T), # Sub-Subskala 2: GenAI Items
  "Ethics" = grep("^TP05_0[1-6]$", names(data), value = T) #im pretest nur items 1-6 vorhanden 7,8 kamen im post dazu
)

# ein paar ethik items müssen noch rausgenommen werden, die nur im posttest erhoben wurden. hier nur check
# data_subset1 <- data %>% select(all_of(unlist(list_items_i_tpack)))
# View(data_subset1)

## Reflection
list_items_ref <- list(
  "Ref" = grep("^RE02_", names(data), value = T)
)
# data_subset3 <- data %>% select(all_of(unlist(list_items_ref)))
# sum(data_subset3 < 0, na.rm = TRUE)

## AI Literacy
# data <- data %>% mutate(AILIT = AI35_02) # rename to sumscore
list_items_ailit <- list(
  "ailit" = grep(
    "^(AI36_04|AI36_10|AI36_11|AI36_13|AI36_15|AI36_16|AI36_17|AI36_19|AI36_24|AI36_25|AI36_27|GL21_01|GL21_02|GL21_03|GL21_04|GL21_05|GL21_06|GL21_07|GL21_08|GL21_09|GL21_10|GL21_11|GL21_12|GL21_13|GL21_14|GL21_15|GL21_16|GL21_17|GL21_18|GL21_19)", # results of all ailit items
    names(data),
    value = T
  ),
  "ailit_horn" = grep(
    "^(AI36_04|AI36_10|AI36_11|AI36_13|AI36_15|AI36_16|AI36_17|AI36_19|AI36_24|AI36_25|AI36_27)",
    names(data),
    value = T
  ), # Hornberger items
  "ailit_glat" = grep(
    "^(GL21_01|GL21_02|GL21_03|GL21_04|GL21_05|GL21_06|GL21_07|GL21_08|GL21_09|GL21_10|GL21_11|GL21_12|GL21_13|GL21_14|GL21_15|GL21_16|GL21_17|GL21_18|GL21_19)",
    names(data),
    value = T
  ) # GLAT items
)

# create new columns with the means of the lists
data <- data %>%
  mutate(
    AILIT = rowSums(cbind(AI35_02, GL20_01), na.rm = TRUE), # default sumscore overall AI Lit
    AILIThorn = rowSums(
      select(., all_of(list_items_ailit$ailit_horn)),
      na.rm = TRUE
    ),
    AILITglat = rowSums(
      select(., all_of(list_items_ailit$ailit_glat)),
      na.rm = TRUE
    ),
    TK = rowMeans(select(., all_of(list_items_i_tpack$TK)), na.rm = TRUE),
    TPK = rowMeans(select(., all_of(list_items_i_tpack$TPK)), na.rm = TRUE),
    TCK = rowMeans(select(., all_of(list_items_i_tpack$TCK)), na.rm = TRUE),
    TPACKce = rowMeans(
      select(., all_of(list_items_i_tpack$TPACK_ce)),
      na.rm = TRUE
    ),
    TPACKgenai = rowMeans(
      select(., all_of(list_items_i_tpack$TPACK_genai)),
      na.rm = TRUE
    ),
    Ethics = rowMeans(
      select(., all_of(list_items_i_tpack$Ethics)),
      na.rm = TRUE
    ),
    Ref = rowMeans(select(., all_of(list_items_ref$Ref)), na.rm = TRUE),
    TPACK_total = rowMeans(
      select(., all_of(unlist(list_items_i_tpack))),
      na.rm = TRUE
    )
  )

# check missing values in TPACK and PU scales (NA means, not answered (-1 diff. answer, -9 not answered))
data %>%
  select(all_of(c(list_items_i_tpack) %>% unlist)) %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  sum # 0

```

# Methods

## Participants

```{r participants}
p_load(dplyr, tidyr) # needs to be called for rendering to work
data_participants <- data %>%
  filter(time == "pre") %>%
  group_by(intervention, location, gender) %>%
  summarise(
    n = n(),
    n_complete = sum(complete.cases(.)),
    age_min = min(age, na.rm = T),
    age_max = max(age, na.rm = T),
    age_mean = mean(age, na.rm = T) %>% round(2),
    age_sd = sd(age, na.rm = T) %>% round(2),
    semester_min = min(semester, na.rm = T),
    semester_max = max(semester, na.rm = T),
    semester_mean = mean(semester, na.rm = T) %>% round(2),
    semester_sd = sd(semester, na.rm = T) %>% round(2)
  ) %>%
  ungroup()
```

### table

```{r table-participants, result = "asis"}
#| label: tbl-participants
#| tbl-cap: "Demographic Data of the Three Experimental Groups."
# render as table
format <- ifelse(
  knitr::is_html_output(),
  "markdown",
  ifelse(knitr::is_latex_output(), "latex", "markdown")
)

var_names <- list(
  gender = "Gender",
  age = "Age",
  semester = "Semester Number",
  students = "Study Level",
  isp = "Completed Integrated Semester Internship",
  subject1 = "Subject 1",
  subject2 = "Subject 2",
  studies = "Study Program",
  location = "University",
  TK = "Intelligent TK",
  TCK = "Intelligent TCK",
  TPK = "Intelligent TPK",
  TPACK_total = "Intelligent TPACK",
  AILIT = "AI Literacy",
  Ethics = "AI Ethics",
  Ref = "Reflection"
)

kable(
  summary(
    tableby(
      intervention ~
        location +
        age +
        gender +
        semester +
        students +
        TK +
        TCK +
        TPK +
        TPACK_total +
        Ethics +
        AILIT +
        Ref,
      data %>%
        filter(time == "pre") %>%
        mutate(
          gender = factor(gender, labels = c("Female", "Male")),
          students = factor(students, labels = c("Bachelor", "Master"))
        ) %>%
        droplevels(),
      cat.test = "chisq",
      chisq.correct = T,
      control = tableby.control(test = T, total = T, digits = 2)
    ),
    labelTranslations = var_names,
    pfootnote = T,
    text = F
  ),
  escape = T,
  format = format,
  booktabs = T
)
```


## Analysis

```{r analysis-setup}
#define vars
controls <- c("gender", "semester", "location")
random_terms_lmer <- c("(1 | code)")
groups <- c("intervention")
time <- "time"
outcomes <- c(
  "AILIT",
  "AILIThorn",
  "AILITglat",
  "TK",
  "TPK",
  "TCK",
  "TPACKce",
  "TPACKgenai",
  "Ethics",
  "Ref"
)
vars_num <- c("age", "semester") # numeric outcomes
vars_cat <- c("gender", "students") # categorical outcomes`
```

## Effect Sizes

```{r effect-sizes}
#| eval: false
lmers_prepost_all <-
  lapply(
    outcomes,
    fit_lmer,
    groups,
    time,
    controls,
    random_terms_lmer,
    NULL,
    data = data
  )

dhat <- function(model, idx = 3, type = 2) {
  f <- abs(tail(summary(model)$coefficients, idx))[1]
  v <- sqrt(tail(anova(model, type = type)[["Mean Sq"]], 1))
  return(f / v)
}
#' Compute standardized effect size d for one fixed effect in an lmer model, based on https://pmc.ncbi.nlm.nih.gov/articles/PMC6646942/
#'
#' d = estimate / sqrt(sum of variances of random effects [+ residual if chosen])
#'
#' @param model lmer model
#' @param effect character, name of the fixed effect in fixef(model)
#' @param include_residual logical, include residual variance (default TRUE)
#' @return numeric, effect size d
lmer_d <- function(model, effect, include_residual = TRUE) {
  stopifnot(effect %in% names(lme4::fixef(model)))

  # numerator
  est <- unname(lme4::fixef(model)[effect])

  # variance components
  vc <- as.data.frame(lme4::VarCorr(model)) # grp, var1, var2, vcov, sdcor
  var_random <- sum(vc$vcov, na.rm = TRUE)

  # residual
  if (!include_residual) {
    var_random <- var_random - sigma(model)^2
  }

  d <- est / sqrt(var_random)
  return(d)
}

effect_sizes_dhat_old <- sapply(lmers_prepost_all, dhat, idx = 2) %>% round(2)

effect_sizes <- sapply(
  lmers_prepost_all,
  lmer_d,
  "intervention-c+STEM_language:time-pre+post"
) %>%
  round(2)
effect_sizes
```


### Ancova
```{r ancova}
# reshape data to wide format for ancova
data_wide <- data %>%
  pivot_wider(
    id_cols = c(
      code,
      intervention,
      group,
      gender,
      semester,
      students,
      location
    ),
    names_from = time,
    values_from = all_of(outcomes)
  ) %>%
  mutate(
    # scale the posttest score so that the estimates are approximations of Cohen's d
    across(all_of(outcomes %>% paste0("_post")), scale),
    treatment = factor(ifelse(
      intervention == "control",
      "control",
      "intervention"
    ))
  )

contrasts.treatment <- cbind(c(-1 / 2, 1 / 2))
colnames(contrasts.treatment) <- c("-c+t")
contrasts(data_wide$treatment) <- contrasts.treatment

ancovas <- lapply(
  outcomes,
  fit_ancova,
  groups,
  # we remove University as a covariate as we only have control students at University 2
  setdiff(controls, "location"),
  "HC3",
  data_wide
)
names(ancovas) <- outcomes

ancovas_two_groups <- lapply(
  outcomes,
  fit_ancova,
  "treatment",
  setdiff(controls, "location"),
  "HC3",
  data_wide
)
names(ancovas_two_groups) <- outcomes

```


## Intelligent TPACK and AI Ethics

::: {.landscape}
```{r tab.layout="autofit"}
#| label: tbl-ancova-tpack-summary
#| tbl-cap: "Summary of Multiple Regression Models for Intelligent TPACK Subscales and AI Ethics"

ft_ancova <- modelsummary_models(
  ancovas[grepl("^(T|Ethics)", names(ancovas))],
  outcomes = outcomes %>% grep("^(T|Ethics)", ., value = TRUE),
  output = modelsummary_output,
  # coef_rename = c("genderMännlich" = "Geschlecht")
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_ancova <- adjust_table_for_word(ft_ancova, width = 1)
}
ft_ancova
```

:::

## AI Literacy

```{r tab.layout = "autofit"}
#| label: tbl-ancova-ailit-summary
#| tbl-cap: "Summary of Multiple Regression Models for AI Literacy."

ft_ailit_ancova <- modelsummary_models(
  ancovas[names(ancovas) %in% c("AILIT", "AILIThorn", "AILITglat")],
  outcomes = c("AILIT", "AILIThorn", "AILITglat"),
  output = modelsummary_output,
  # coef_rename = c("genderMännlich" = "Geschlecht")
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_ailit_ancova <- adjust_table_for_word(ft_ailit_ancova, width = .75)
}
ft_ailit_ancova
```

## Critical Reflection

::: {.landscape}
```{r tab.layout="autofit"}
#| label: tbl-ancova-ref-summary
#| tbl-cap: "Summary of Multiple Regression Models for Reflection subscale."

ft_ailit_ancova <- modelsummary_models(
  ancovas[names(ancovas) %in% c("Ref")],
  outcomes = c("Ref"),
  output = modelsummary_output,
  # coef_rename = c("genderMännlich" = "Geschlecht")
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_ailit_ancova <- adjust_table_for_word(ft_ailit_ancova, width = .75)
}
ft_ailit_ancova
```

:::

```{r, results="asis"}
if (!is_word) {
  cat("### Pairwise Contrasts Between Groups")
}
```

```{r ailit-ancova-between-contrasts}
# pairwise between contrasts (no multiplicity adjustment)
ailit_ancova_between_contrasts <- lapply(ancovas[1:3], function(m) {
  emmeans(m, ~intervention) %>%
    contrast(interaction = "pairwise", adjust = "none")
})
if (!is_word) {
  print(ailit_ancova_between_contrasts)
}
```

## Two Groups (Control vs. Intervention)

::: {.callout-note}
Due to the very small (insignificant) difference between the distribution of gender in the treatment groups, we could drop it from the covariates.

:::

```{r effectsizes-covariates-two-groups}
#| eval: false
# perform chi square for categorical variables
chisquare_covariates_two_groups <- vars_cat %>%
  lapply(function(v) {
    tab <- table(d[[v]], d$treatment)
    chisq <- chisq.test(tab)
  })
names(chisquare_covariates_two_groups) <- vars_cat

chisquare_covariates_two_groups

effectsizes_covariates__two_groups %>% kable
```

```{r ancova-two-groups}
#| label: tbl-ancova-two-groups
#| tbl-cap: "Summary of Multiple Linear Regression Models for Two Treatment Groups."

modelsummary_models(
  ancovas_two_groups,
  outcomes = outcomes,
  output = modelsummary_output
)

```


```{r save-env }
# save environment
save.image("data/env-analysis.Rdata")
```