---
title: "WoLKE-Pilot-2025"
author: "Benedikt Beuttler"
date: "`r Sys.Date()`"
lang: de
page-layout: full
document-css: false
link-citations: true
number-sections: true
editor: visual
format:
  html:
    standalone: true
    embed-resources: true
    section-divs: true
    html-math-method: mathjax
    wrap: none
    default-image-extension: png
    date-format: long
---

# Preamble

## Prepare the Environment

## Load Libraries

```{r load-libraries-and-functions}
#| output: false
library(this.path)
library(dplyr) #e.g. for select
library(tidyr) # for reshaping data
library(stringdist) #levenshtein
library(stringr) # for reshaping data
library(lme4)
library(lmerTest)
library(emmeans)
library(modelsummary)
library(kableExtra)
library(arsenal)
library(forcats) 
source(file="functions.r") # all outsourced funtions
```

## Load Pre- and Posttest and Reshape Data

```{r load-data}
# Set working directory to the script's location
setwd(this.path::this.dir())
### PRETEST
# load csv -> df_pre
source("./data/import_ki-lehramt_2025-07-24_08-36.r")

# extract items starting with TP (TPACK) or DE (Demographics)
data_pre <- df_pre %>% select(QUESTNNR, starts_with("DE"), starts_with("AI"), starts_with("ED"), starts_with("TP"), starts_with("RE"), starts_with("CT"), starts_with("STARTED"), starts_with("FINISHED"))
data_pre$QUESTNNR[substr(data_pre$STARTED, 6, 7) == "07"] <- "mint-post" # fixe some wrong entries in pre list to post

# rename demographics
data_pre <- data_pre %>%
  mutate(
    code       = tolower(DE13_01),
    age        = DE14_01,
    semester   = DE15_01,
    studies    = DE16,
    subject1   = DE24,
    subject2   = DE25,
    gender     = DE10,
    location   = DE11,
    semestercount = DE15_01,
    isp        = DE09,
    #sprache_intervention = DE08,
    sprache_last_sem = DE06, #already visited cans KI seminar
    #mint_control = DE07,
    mint_exclusion = DE02,
    intervention_sprache = case_when(
      str_to_lower(as.character(DE08)) == "ja" ~ TRUE,
      str_to_lower(as.character(DE08)) == "nein" ~ FALSE
    ),
    intervention_mint = case_when(
      str_to_lower(as.character(DE07)) == "ja" ~ TRUE,
      str_to_lower(as.character(DE07)) == "nein" ~ FALSE
    ),
    intervention_only = coalesce(intervention_sprache, FALSE) | coalesce(intervention_mint, FALSE),
    intervention_only = as.logical(intervention_only)
  )
data_pre <- data_pre[, c("code", setdiff(names(data_pre), "code"))] #set code as first column
# View(data_pre[(is.na(data_pre$code) | data_pre$code == ""), ])
data_pre <- data_pre[!(is.na(data_pre$code) | data_pre$code == ""), ] # remove empty/na codes (=7) -> 1 stimmt Datenschutz nicht zu

#### POSTTEST

# load csv -> df_post
source("./data/import_ki-lehramt_2025-07-24_08-35.r")

# extract items starting with
data_post <- df_post %>% select(QUESTNNR, starts_with("DE"), starts_with("AI"), starts_with("ED"), starts_with("TP"), starts_with("RE"), starts_with("CT"), starts_with("STARTED"), starts_with("FINISHED"))

# rename demographics
data_post <- data_post %>%
  mutate(
    code       = tolower(DE13_01),
    age        = DE14_01,
    semester   = DE15_01,
    studies    = DE16,
    subject1   = DE24,
    subject2   = DE25,
    gender     = DE10,
    location   = DE11,
    semestercount = DE15_01,
    isp        = DE09,
    # sprache_intervention = DE08,
    #sprache_info = DE06, #already visited cans KI seminar
    # mint_control = DE07,
    mint_exclusion = DE02,
    intervention_sprache = case_when(
      str_to_lower(as.character(DE08)) == "ja" ~ TRUE,
      str_to_lower(as.character(DE08)) == "nein" ~ FALSE
    ),
    intervention_mint = case_when(
      str_to_lower(as.character(DE07)) == "ja" ~ TRUE,
      str_to_lower(as.character(DE07)) == "nein" ~ FALSE
    ),
    intervention_only = coalesce(intervention_sprache, FALSE) | coalesce(intervention_mint, FALSE),
    intervention_only = as.logical(intervention_only)
  )
data_post <- data_post[, c("code", setdiff(names(data_post), "code"))] #set code as first column
# View(data_post[(is.na(data_post$code) | data_post$code == ""), ])
data_post <- data_post[!(is.na(data_post$code) | data_post$code == ""), ] # remove empty/na codes (=8) -> 1 hat schon mint besucht, 2 stimmen nicht Datenschutz zu

# remove non basic types to make it mergeable
# Convert everything to character, then use type.convert to infer proper base types
clean_df <- function(df) {
  df[] <- lapply(df, function(col) {
    if (!is.atomic(col)) col <- as.character(col)
    col <- as.character(col)  # force avector etc. to become characters
    type.convert(col, as.is = TRUE)
  })
  return(df)
}
data_pre_clean <- clean_df(data_pre)
data_post_clean <- clean_df(data_post)

# merge pre und post dfs
#data <- dplyr::bind_rows(data_pre, data_post)
data <- bind_rows(data_pre_clean, data_post_clean)

# CT-Score aus CT28_01 übernehmen (-1/-9 als NA) und als CT_total_mean setzen
data <- data %>%
  mutate(
    CT28_01 = suppressWarnings(as.numeric(CT28_01)),
    CT28_01 = ifelse(CT28_01 %in% c(-1, -9), NA, CT28_01),
    CT_total_mean = CT28_01
  )

# factorize some columns
data$gender <- factor(data$gender)
data$location <- factor(data$location)

# create relevant columns group (sprache mint), time (pre, post) and intervention (control, mint, sprache)
data <- data %>%
  mutate(
    group = case_when(
      str_detect(QUESTNNR, "sprache") ~ "sprache",
      str_detect(QUESTNNR, "mint") ~ "mint"
    ),
    time = case_when(
      str_detect(QUESTNNR, "post") ~ "post",
      TRUE ~ "pre"
    ),
    group = factor(group),
    time = factor(time, levels=c("pre","post"))
  )

data <- data %>% mutate(
  intervention = factor(ifelse(intervention_only == F, "control",
  ifelse(group == "mint", "mint", "sprache")))
)

### EXCLUSIONS
e1 <- nrow(data[data$FINISHED == "FALSE", ]) # not FINISHED
e1
data <- data[data$FINISHED == "TRUE", ] 

e2 <- nrow(data[data$code == "höel07", ])
e2
data <- data[data$code != "höel07", ] #example code

# check for multiple entries and completed questionnaires
# view a specific code; obsu02 stfr08 (both multiple finished questionnaire, rowindex 1811)
data <- data[!rownames(data) %in% c("1811", "1814"), ]
# View(data[data$code == "obsu02", ])

# Code corrections from levenshtein checks
data$code[data$code == "mair10"] <- "moir10"
data$code[data$code == "drch17"] <- "frch17"
data$code[data$code == "loma26"] <- "goma26"
data$code[data$code == "sisu00"] <- "sisu01"
data$code[data$code == "mosy24"] <- "stsy24"
data$code[data$code == "frco30"] <- "scco30"
data$code[data$code == "meli06"] <- "meli63"
data$code[data$code == "waol30"] <- "spol30"

# check code distances
check_distances <- F
if (check_distances){
  # finding typos in code
  # get and save single instances of code
  single_codes <- data %>%
    count(code) %>%
    filter(n == 1) %>%
    pull(code)
  # Get only the rows where code is in single_codes
  data_single <- data %>%
    filter(code %in% single_codes)
  
  # Now split the codes based on group
  single_codes_pre <- data_single %>%
    filter(time == "pre") %>%
    pull(code)
  
  single_codes_post <- data_single %>%
    filter(time == "post") %>%
    pull(code)
  
  # 1. Compute the distance matrix
  dist_matrix <- stringdistmatrix(a = single_codes_pre, b = single_codes_post, method = "lv")
  
  # 2. Convert to data frame with row and column names
  dist_df <- as.data.frame(dist_matrix)
  rownames(dist_df) <- single_codes_pre
  colnames(dist_df) <- single_codes_post
  
  # 3. Extract the index positions where distance == 2
  matches_2 <- which(dist_df == 2, arr.ind = TRUE)
  
  # 4. Create a data frame of those matches
  matches_distance_2 <- data.frame(
    pre_code = rownames(dist_df)[matches_2[, "row"]],
    post_code = colnames(dist_df)[matches_2[, "col"]],
    distance = 2
  )
  
  # 5. View result
  # View(matches_distance_2)
}

# # create new columns with scale naming
# data <- data %>%
#   mutate(across(starts_with("TP01_"), identity, .names = "TPACK_TK_{str_sub(.col, -2)}")) %>%
#   mutate(across(starts_with("TP02_"), identity, .names = "TPACK_TPK_{str_sub(.col, -2)}")) %>%
#   mutate(across(starts_with("TP03_"), identity, .names = "TPACK_TCK_{str_sub(.col, -2)}")) %>%
#   mutate(across(starts_with("TP04_"), identity, .names = "TPACK_TPACK_{str_sub(.col, -2)}")) %>%
#   mutate(across(starts_with("TP05_"), identity, .names = "TPACK_Ethic_{str_sub(.col, -2)}"))

```

```{r Demographics}
# Generate the tableby summary
preprocess_data <- function(data) {
  data %>%
    droplevels() %>%
    mutate(
      studies  = fct_infreq(studies),
      subject1 = fct_infreq(subject1),
      subject2 = fct_infreq(subject2)
    )
}

# only get data from pretest for demographic data to see registered
data_pretest <- data %>% 
  filter(time == "pre")
nrow(data_pretest)

# only take samples with pre and post test data into account, singlecodes
data_dem_all <- data %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  filter(time == "post") %>%
  ungroup()
nrow(data_dem_all)

# Create the formula
demographics_formula <- intervention ~  location + age  + gender + semester + isp + studies + subject1 + subject2
demographics <- tableby(demographics_formula, data = data_dem_all)

# Ausgabe als Tabellen
summary(demographics) %>%
  as.data.frame() %>%
  knitr::kable(caption = "Demographic Summary")

# ---------

# # 1. Sprache
# data_sprache <- data_pretest %>%
#   filter(group == "sprache") %>%
#   preprocess_data()

# demographics_sprache <- tableby(demographics_formula, data = data_sprache)

# # 2. MINT Intervention
# data_mint_intervention <- data_pretest %>%
#   filter(group == "mint", intervention_only == TRUE) %>%
#   preprocess_data()

# demographics_mint_intervention <- tableby(demographics_formula, data = data_mint_intervention)

# # 3. MINT Kontrollgruppe
# data_mint_control <- data_pretest %>%
#   filter(group == "mint", intervention_only == FALSE) %>%
#   preprocess_data()

# demographics_mint_control <- tableby(demographics_formula, data = data_mint_control)

# # Ausgabe als Tabellen
# summary(demographics_sprache) %>%
#   as.data.frame() %>%
#   knitr::kable(caption = "Demographic Summary: Sprache-Gruppe")

# summary(demographics_mint_intervention) %>%
#   as.data.frame() %>%
#   knitr::kable(caption = "Demographic Summary: MINT Intervention")

# summary(demographics_mint_control) %>%
#   as.data.frame() %>%
#   knitr::kable(caption = "Demographic Summary: MINT Kontrollgruppe")

# TODO: only n's of both timepoints

```

```{r Prepare scales}

### grouping scales
## TPACK
# all TPACK items used
list_items_i_tpack <- list(
  "TK" = grep("^TP01_", names(data), value=T),
  "TPK" = grep("^TP02_", names(data), value=T),
  "TCK" = grep("^TP03_", names(data), value=T),
  "TPACK_appl" = grep("^TP04_(01|02|03|09|10|11|12|13|14)$", names(data), value=T), # Sub-Subskala 1: Anwendung
  "TPACK_intrg" = grep("^TP04_(04|05|06|07|08|15)$", names(data), value=T), # Sub-Subskala 2: Integration
  "TPACK_genai" = grep("^TP04_(09|10|11|13|14)$", names(data), value=T), # Sub-Subskala 3: GenAI Items
  "Ethics" = grep("^TP05_0[1-6]$", names(data), value=T) #im pretest nur items 1-6 vorhanden 7,8 kamen im post dazu
)

# ein paar ethik items müssen noch rausgenommen werden, die nur im posttest erhoben wurden. hier nur check
# data_subset1 <- data %>% select(all_of(unlist(list_items_i_tpack)))
# View(data_subset1)

## Perceived Usefulnes
list_items_pu <- list(
  "Pu" = grep("^ED03_", names(data), value=T)
)
# data_subset2 <- data %>% select(all_of(unlist(list_items_pu)))
# sum(data_subset2 < 0, na.rm = TRUE)

## Reflection (bei Piloten irrelevant)
# list_items_ref <- list(
#   "Ref" = grep("^RE01_", names(data), value=T)
# )
# data_subset3 <- data %>% select(all_of(unlist(list_items_ref)))
# sum(data_subset3 < 0, na.rm = TRUE)

## AI Literacy
# TODO: do we want to split the results for each subscale?
# data <- data %>% mutate(AILIT = AI35_02) # rename to sumscore

# create new columns with the means of the lists
data <- data %>%
  mutate(
    AILIT = AI35_02,
    TK     = rowMeans(select(., all_of(list_items_i_tpack$TK)), na.rm = TRUE),
    TPK    = rowMeans(select(., all_of(list_items_i_tpack$TPK)), na.rm = TRUE),
    TCK    = rowMeans(select(., all_of(list_items_i_tpack$TCK)), na.rm = TRUE),
    TPACKappl  = rowMeans(select(., all_of(list_items_i_tpack$TPACK_appl)), na.rm = TRUE),
    TPACKintrg  = rowMeans(select(., all_of(list_items_i_tpack$TPACK_intrg)), na.rm = TRUE),
    TPACKintrg  = rowMeans(select(., all_of(list_items_i_tpack$TPACK_intrg)), na.rm = TRUE),
    TPACKgenai  = rowMeans(select(., all_of(list_items_i_tpack$TPACK_genai)), na.rm = TRUE),
    Ethics = rowMeans(select(., all_of(list_items_i_tpack$Ethics)), na.rm = TRUE),
    PU     = rowMeans(select(., all_of(list_items_pu$Pu)), na.rm = TRUE),
    # Ref    = rowMeans(select(., all_of(list_items_ref$Ref)), na.rm = TRUE),
    TPACK_total = rowMeans(select(., all_of(unlist(list_items_i_tpack))), na.rm = TRUE)
  )

```

```{r lmer-Sprache}

#### SPRACHE ####
# nur daten berücksichtigen die pre und post test haben
data_sprache <- data %>%
  filter(group == "sprache") %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  ungroup()

## define variables
random_terms_lmer <- c("(1 | code)")
# groups <- c("intervention") 
outcomes <- c("AILIT", "TK", "TPK", "TCK", "TPACKappl", "TPACKintrg", "TPACKgenai", "Ethics", "PU")
time <- "time"
controls <- c("age", "gender")

# groups set as NA
lmers_prepost <-
  lapply(outcomes, fit_lmer, NULL, time, controls, random_terms_lmer, NULL, data= data_sprache)

modelsummary_models(lmers_prepost)
```

Im folgenden verwenden wir drei Kontraste: (0) Intercept = Gesamtdurchschnitt aller Gruppen, (1) Exp-Ø (MINT+Sprache)/2 vs. Control, (2) Sprache vs. MINT. Mit der Zeit-Interaktion testen wir, ob sich diese Unterschiede über die Zeit unterschiedlich entwickeln (Difference-in-Differences: Exp-Ø vs. Control sowie Sprache vs. MINT).
``` {r lmer MINT vs Sprache vs Kontrollgruppe}
data_all <- data %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  ungroup()

## define contrast Control vs Exp1+Exp2 - vermutlich bessere version, da die Estimates "stimmen"
mat <- cbind(c(1 / 3, 1 / 3, 1 / 3), # Durchschnitt aller 3 Gruppen
  c(-1, .5, .5),                     # Vergleich Control vs Mittel der beiden Interventions-Gruppen
  c(0, -1, 1))                       # Vergleich Sprache vs MINT
contrasts.CvsExp <- solve(t(mat))[, -1]
colnames(contrasts.CvsExp) <- c("-C+MINT_Sprache", "-MINT+Sprache")
contrasts(data_all$intervention) <- contrasts.CvsExp

# or use this coding system?
# my.helmert <- matrix(c(-2 / 3, 1 / 3, 1 / 3, 0, -1 / 3, 1 / 3), ncol = 2)
# colnames(my.helmert) <- c("-C+MINT_Sprache", "-MINT+Sprache")
# contrasts(data$cond2) <- my.helmert

## define variables
random_terms_lmer <- c("(1 | code)")
groups <- c("intervention")
outcomes <- c("AILIT", "TK", "TPK", "TCK", "TPACKappl", "TPACKintrg", "TPACKgenai", "Ethics", "PU")
time <- "time"
controls <- c("age", "gender", "location")

# groups set as NA
lmers_prepost_all <-
  lapply(outcomes, fit_lmer, groups, time, controls, random_terms_lmer, NULL, data = data_all)

modelsummary_models(lmers_prepost_all)


# ---------- klappt nich lol
data_sg <- data %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  filter(location=="PH Schwäbisch Gmünd") %>%
  ungroup()
controls <- c("age", "gender")

lmers_prepost_sg <-
  lapply(outcomes, lmer, NULL, time, controls, random_terms_lmer, NULL, data = data_sg)

modelsummary_models(lmers_prepost_sg)



# mm <- lmer(PU ~ time*intervention + age + gender + (1 | code), data = data %>% filter(all(c("pre", "post") %in% time)))
# mm %>% summary(corr=F)
# emmeans(lmers_prepost_all[[1]], ~ intervention*time|location, lmer.df = "satterthwaite") %>% contrast(interaction = "pairwise", adjust="none")
```

``` {r old code}

# Pre/Post-Effekt auf CT
model_ct_sprache <- lmer(
  CT_total_mean ~ time + age + gender + (1 | code),
  data = data_sprache
)
summary(model_ct_sprache)

# Deskriptiv: CT nach Zeit (Sprache)
ct_desc_sprache <- data_sprache %>%
  group_by(time) %>%
  summarise(
    n    = sum(!is.na(CT_total_mean)),
    mean = mean(CT_total_mean, na.rm = TRUE),
    sd   = sd(CT_total_mean,   na.rm = TRUE),
    .groups = "drop"
  )
knitr::kable(ct_desc_sprache, caption = "CT (Sprache): Deskriptive Statistiken Pre/Post")


#### MINT - nur SG ####
# 1. MINT-Daten mit Pre- und Posttest
data_mint <- data %>%
  filter(group == "mint") %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  ungroup()
# 2. Subset nur für PH Schwäbisch Gmünd
data_mint_sg <- data_mint %>%
  filter(location == "PH Schwäbisch Gmünd")

# 3. AI lit Sumscore mit Intervention vs. Kontrollgruppe, nur für Schwäbisch Gmünd
model_ailit_contr_mint_sg <- lmer(AILIT ~ time * intervention + age + gender + (1 | code), data = data_mint_sg)
summary(model_ailit_contr_mint_sg)

# TPACK-Modelle nur für Schwäbisch Gmünd: Zeit (pre vs. post)
model_tpack_mint_sg <- lmer(TPACK_total ~ time + age + gender + (1 | code), data = data_mint_sg)
summary(model_tpack_mint_sg)

# Interaktion: Zeit * Intervention
model_tpack_int_contr_mint_sg <- lmer(TPACK_total ~ time * intervention + age + gender + (1 | code), data = data_mint_sg)
# summary(model_tpack_contr_mint_sg) TODO: fix Error in h(simpleError(msg, call)) :  error in evaluating the argument 'object' in selecting a method for function 'summary': object 'model_tpack_contr_mint_sg' not found


#### MINT ####
# data_mint <- data %>% filter(group == "mint")
data_mint <- data %>%
  filter(group == "mint") %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  ungroup()

# AI Lit Sumscore
model_ailit_mint <- lmer(AILIT ~ time + age + gender + (1 | code), data = data_mint)
summary(model_ailit_mint) # sign.
model_ailit_loc_mint <- lmer(AILIT ~ time * location + age + gender + (1 | code), data = data_mint)
summary(model_ailit_loc_mint) # no sign. diff between phsg und phlb
model_ailit_contr_mint <- lmer(AILIT ~ time * intervention + age + gender + (1 | code), data = data_mint)
summary(model_ailit_contr_mint) # höherer posttest score als zum pretest, aber nicht verglichen der gruppen über zeit: intervention ist nicht unbedingt der grund warum die scores höher sind

# Linear mixed effects model: TPACK_total ~ time (fixed effect), with random intercept per person (code)
model_tpack_mint <- lmer(TPACK_total ~ time + age + gender + (1 | code), data = data_mint)
summary(model_tpack_mint) # sign.
model_tpack_loc_mint <- lmer(TPACK_total ~ time * location + age + gender + (1 | code), data = data_mint)
summary(model_tpack_loc_mint) # sign. PH Schwäbisch Gmünd shows a smaller improvement -> maybe cause of control?
model_tpack_contr_mint <- lmer(TPACK_total ~ time * intervention + age + gender + (1 | code), data = data_mint)
summary(model_tpack_contr_mint) # control group did not improve, intervention improved sign. from pre to post (see interaction)

# Perceived Usefulness
model_pu_mint <- lmer(PU ~ time + age + gender + (1 | code), data = data_mint)
summary(model_pu_mint) # no sign.


#### CT (Computational Thinking) – MINT ####

# Gesamt MINT: Haupteffekt Zeit
model_ct_time_mint <- lmer(
  CT_total_mean ~ time + age + gender + (1 | code),
  data = data_mint
)
summary(model_ct_time_mint)

# Gesamt MINT: Zeit * Intervention
model_ct_timeXint_mint <- lmer(
  CT_total_mean ~ time * intervention + age + gender + (1 | code),
  data = data_mint
)
summary(model_ct_timeXint_mint)

# Gesamt MINT: Zeit * Location
model_ct_timeXloc_mint <- lmer(
  CT_total_mean ~ time * location + age + gender + (1 | code),
  data = data_mint
)
summary(model_ct_timeXloc_mint)

# Nur PH Schwäbisch Gmünd
data_mint_sg <- data_mint %>% filter(location == "PH Schwäbisch Gmünd")

model_ct_time_mint_sg <- lmer(
  CT_total_mean ~ time + age + gender + (1 | code),
  data = data_mint_sg
)
summary(model_ct_time_mint_sg)

model_ct_timeXint_mint_sg <- lmer(
  CT_total_mean ~ time * intervention + age + gender + (1 | code),
  data = data_mint_sg
)
summary(model_ct_timeXint_mint_sg)

```