---
title: "WoLKE-Pilot-2025"
author: "Benedikt Beuttler"
date: "`r Sys.Date()`"
lang: de
page-layout: full
document-css: false
link-citations: true
number-sections: true
toc: true
csl: https://www.zotero.org/styles/apa
bibliography: library.bib
editor: source
execute:
  cache: true
format:
  html:
    standalone: true
    embed-resources: true
    section-divs: true
    html-math-method: mathjax
    wrap: none
    default-image-extension: png
    date-format: long
---

# Preamble

## Prepare the Environment

## Load Libraries

```{r load-libraries-and-functions}
#| output: false
library(this.path)
library(dplyr) #e.g. for select
library(tidyr) # for reshaping data
library(stringdist) #levenshtein
library(stringr) # for reshaping data
library(lme4)
library(lmerTest)
library(emmeans)
library(modelsummary)
library(kableExtra)
library(arsenal)
library(forcats)
library(psych)
library(purrr)
library(ggplot2)
library(ggsci) # color palettes
library(jtools) # theme_apa etc.
source(file = "functions.r") # all outsourced funtions
```

## Load Pre- and Posttest and Reshape Data

```{r load-data}
# Set working directory to the script's location
setwd(this.path::this.dir())
### Load Data
# load csv -> df_pre
source("./data/import_ki-lehramt.r")

# View(df[tolower(df$DE13_01) == "stfr08", ]) #stfr08

# extract items starting with TP (TPACK) or DE (Demographics)
data_raw <- df %>%
  select(
    QUESTNNR,
    starts_with("DE"),
    starts_with("AI"),
    starts_with("ED"),
    starts_with("TP"),
    starts_with("RE"),
    starts_with("CT"),
    starts_with("STARTED"),
    starts_with("FINISHED")
  )
# sum(data_raw$QUESTNNR == "mint-post") # 10 are incorrectly labeled
data_raw$QUESTNNR[
  substr(data_raw$STARTED, 6, 7) == "07" &
    grepl("mint", data_raw$QUESTNNR)
] <- "mint-post" # fixe some wrong entries in pre list to post

# after checking dem_diff later, we see subjs with different group/location values in pre and post (döne15, HONA18)
# correct them here as early as possible
data_raw$QUESTNNR[!is.na(data_raw$DE13_01) & data_raw$DE13_01 == "DÖNE15"][
  1
] <- "sprache"
data_raw$DE08[!is.na(data_raw$DE13_01) & data_raw$DE13_01 == "DÖNE15"][
  1
] <- "Ja"
data_raw$DE11[!is.na(data_raw$DE13_01) & data_raw$DE13_01 == "HONA18"][
  1
] <- "PH Schwäbisch Gmünd"

# rename demographics
data_raw <- data_raw %>%
  mutate(
    code = tolower(DE13_01),
    age = DE14_01,
    semester = DE15_01,
    studies = DE16,
    subject1 = DE24,
    subject2 = DE25,
    gender = DE10,
    location = DE11,
    semestercount = DE15_01,
    isp = DE09,
    #sprache_intervention = DE08,
    sprache_last_sem = DE06, #already visited cans KI seminar
    #mint_control = DE07,
    mint_exclusion = DE02,
    intervention_sprache = case_when(
      str_to_lower(as.character(DE08)) == "ja" ~ TRUE,
      str_to_lower(as.character(DE08)) == "nein" ~ FALSE
    ),
    intervention_mint = case_when(
      str_to_lower(as.character(DE07)) == "ja" ~ TRUE,
      str_to_lower(as.character(DE07)) == "nein" ~ FALSE
    ),
    intervention_only = coalesce(intervention_sprache, FALSE) |
      coalesce(intervention_mint, FALSE),
    intervention_only = as.logical(intervention_only)
  )
data_raw <- data_raw[, c("code", setdiff(names(data_raw), "code"))] #set code as first column
# View(data_raw[(is.na(data_raw$code) | data_raw$code == ""), ])
data_raw <- data_raw[!(is.na(data_raw$code) | data_raw$code == ""), ] # remove empty/na codes (=27)

data_raw_clean <- clean_df(data_raw)
# data_raw_clean <- data_raw

## factorize some columns
# define new levels for gender:
# levels(data_raw_clean$gender) <- c("diverse", "no answer", "male", "female")
# change order of levels
data_raw_clean$gender <- factor(
  data_raw_clean$gender
  # levels = c("female", "male", "diverse", "no answer")
)

# for location
data_raw_clean$location <- factor(
  data_raw_clean$location,
  levels = c("PH Ludwigsburg", "PH Schwäbisch Gmünd")
)

# identify columns of interest
cols_to_check <- grep(
  "^(CT|TP|AI36|ED|RE)",
  names(data_raw_clean),
  value = TRUE
)

# reshape to long format and check for negatives
negatives_df <- data_raw_clean %>%
  select(code, all_of(cols_to_check)) %>% # keep ID/code and relevant columns
  pivot_longer(
    cols = all_of(cols_to_check),
    names_to = "variable",
    values_to = "value"
  ) %>%
  filter(value < 0)

# list of codes where a negative was found
codes_with_negatives <- unique(negatives_df$code)

# count how many negatives per code
counts_per_code <- negatives_df %>%
  count(code, name = "n_negatives")

# total number of negative instances
total_negatives <- nrow(negatives_df)

# outputs:
# codes_with_negatives # vector of codes
# counts_per_code # table of how many negatives per code
# total_negatives # overall number of negative values

# replace negatives with NA in the original dataframe
data_raw_clean <- data_raw_clean %>%
  mutate(across(all_of(cols_to_check), ~ ifelse(. < 0, NA, .)))

# merge pre und post dfs
data <- data_raw_clean

# create relevant columns group (sprache mint), time (pre, post) and intervention (control, mint, sprache)
data <- data %>%
  mutate(
    group = case_when(
      str_detect(QUESTNNR, "sprache") ~ "language",
      str_detect(QUESTNNR, "mint") ~ "stem"
    ),
    time = case_when(
      str_detect(QUESTNNR, "post") ~ "post",
      TRUE ~ "pre"
    ),
    group = factor(group, levels = c("language", "stem")),
    time = factor(time, levels = c("pre", "post"))
  )

# create intervention with 3 factors: control, mint, sprache
data <- data %>%
  mutate(
    intervention = factor(
      ifelse(
        intervention_only == F,
        "control",
        ifelse(group == "stem", "stem", "language")
      ),
      levels = c("control", "stem", "language")
    )
  )

### EXCLUSIONS
e1 <- nrow(data[data$FINISHED == F, ]) # not FINISHED
e1
data <- data[data$FINISHED == T, ] # only keep completed questionnaires

e2 <- nrow(data[data$code == "höel07", ])
e2
data <- data[data$code != "höel07", ] #example code

# check for multiple entries and completed questionnaires
# table(table(data$code))
# view a specific code; obsu02 stfr08 (both multiple finished questionnaire, rowindex 1811)
# rownames(data)[data$code == "stfr08"]
data <- data[!rownames(data) %in% c("1811", "1814"), ] # TODO: Wie haben wir uns entschieden? am besten erster eintrag nehmen
# View(data_raw[data_raw$code == "obsu02", ]) #stfr08

## Corrections

# Code corrections from levenshtein checks
data$code[data$code == "mair10"] <- "moir10"
data$code[data$code == "drch17"] <- "frch17"
data$code[data$code == "loma26"] <- "goma26"
data$code[data$code == "sisu00"] <- "sisu01"
data$code[data$code == "mosy24"] <- "stsy24"
data$code[data$code == "frco30"] <- "scco30"
data$code[data$code == "meli06"] <- "meli63"
data$code[data$code == "waol30"] <- "spol30"

# investigate: df with subjects having unclear studies, check semestercount to see if BA or MA
temp_df <- data %>%
  filter(!is.na(DE16_16)) %>%
  select(code, DE16_16, semester, intervention)

# append (Master) to those with semestercount > 6 according to tmp_df
data$DE16_16[data$code %in% temp_df$code] <-
  paste0(data$DE16_16[data$code %in% temp_df$code], " (Master)")

remove(temp_df)

# create new variable subject with levels bachelor and master depending on studies value
data <- data %>%
  mutate(
    students = case_when(
      str_detect(tolower(studies), "master") ~ "master",
      str_detect(tolower(studies), "bachelor") ~ "bachelor",
      str_detect(tolower(studies), "nachstudium") ~ "master", # hier gibt es eine person, die nachstudium angegeben hat, aber mit semestercount 6/8 setzen wir ihn auf master
      str_detect(tolower(studies), "sonstiges") &
        str_detect(tolower(DE16_16), "master") ~
        "master",
      str_detect(tolower(studies), "sonstiges") &
        str_detect(tolower(DE16_16), "bachelor") ~
        "bachelor",
      TRUE ~ NA
    ),
    students = factor(students, levels = c("bachelor", "master"))
  )

# check code distances
check_dist <- F
if (check_dist) {
  check_distances(data, 2)
}

# check demograph. differences pre to post
dem_diff <- check_demo_vars(
  data,
  demo_vars = c(
    "age",
    "semester",
    "gender",
    "students",
    "intervention",
    "location"
  ),
  id = "code",
  show_only_diff = T
)

codes_age <- dem_diff %>%
  filter(variable == "age") %>%
  pull(code)
# correct age in post to age in pre (to report one age and have same age for subj when calc models)
for (cc in codes_age) {
  # save pre value
  pre_val <- data$age[data$code == cc & data$time == "pre"]

  # write that in post value
  data$age[data$code == cc & data$time == "post"] <- pre_val
}

codes_gender <- dem_diff %>%
  filter(variable == "gender") %>%
  pull(code)
# correct gender in post to gender in pre
for (cc in codes_gender) {
  # save pre value
  pre_val <- data$gender[data$code == cc & data$time == "pre"]

  # write that in post value
  data$gender[data$code == cc & data$time == "post"] <- pre_val
}

# look at data entries for codes kesü16, leel08, mail19, obpe30, osel08, rosa20, röma24, scna18, stfr08
# codes_of_interest <- c(
#   "rosa20"
# )
# df_subset <- data %>%
#   filter(code %in% codes_of_interest) %>%
#   select(
#     code,
#     subject1,
#     subject2,
#     students,
#     studies,
#     gender,
#     age,
#     location,
#     semester,
#     time,
#     intervention,
#     isp
#   )
# View(df_subset)

# semester count correction (post to pre) after reviewing dem. data
adjust_semester_to_pre <- c(
  "rosa20",
  "luil07",
  "mije10",
  "scco30",
  "leel08",
  "scna18",
  "röma24",
  "osel08"
)
for (cc in adjust_semester_to_pre) {
  # save pre value
  pre_val <- data$semester[data$code == cc & data$time == "pre"]

  # write that in post value
  data$semester[data$code == cc & data$time == "post"] <- pre_val
}

# semester count correction (pre to post) after reviewing dem. data
adjust_semester_to_post <- c("kesü16", "stfr08", "obpe30", "mail19")
for (cc in adjust_semester_to_post) {
  # save pre value
  pre_val <- data$semester[data$code == cc & data$time == "post"]

  # write that in post value
  data$semester[data$code == cc & data$time == "pre"] <- pre_val
}

data_all <- data
# save data for demographics (no pre-post filtering)
save(data_all, file = "data/data_all.Rdata")

# only keep subjects with both data in pre and post test for analysis
data <- data %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  ungroup()

save(data, file = "data/data.Rdata")

```

```{r Demographics}
# Generate the tableby summary
preprocess_data <- function(data_all) {
  data %>%
    droplevels() %>%
    mutate(
      studies = fct_infreq(studies),
      subject1 = fct_infreq(subject1),
      subject2 = fct_infreq(subject2)
    )
}

# only get data from pretest for demographic data to see registered
data_pretest <- data_all %>%
  filter(time == "pre")
nrow(data_pretest) # amount of registered participants 145

# only take samples with pre and post test data into account, singlecodes
data_dem_all <- data_all %>%
  group_by(code) %>%
  filter(all(c("pre", "post") %in% time)) %>%
  filter(time == "post") %>%
  ungroup()
nrow(data_dem_all) # 76 participants with pre and post test

# Create the formula
demographics_formula <- intervention ~
  location + age + gender + semester + isp + studies #+ subject1 + subject2
demographics <- tableby(demographics_formula, data = data_dem_all)

# Ausgabe als Tabellen
summary(demographics) %>%
  as.data.frame() %>%
  knitr::kable(caption = "Demographic Summary")

# Welche Fächer gibt es wie oft?
subject_counts <- data_dem_all %>%
  group_by(intervention) %>%
  pivot_longer(
    c(subject1, subject2),
    names_to = "slot",
    values_to = "subject"
  ) %>%
  mutate(subject = str_squish(subject)) %>%
  filter(!is.na(subject), subject != "") %>%
  count(subject, sort = TRUE, name = "n") #%>%
# mutate(p = n / sum(n))

print(n = 100, subject_counts)

```

```{r Prepare scales}

### grouping scales
## TPACK
# all TPACK items used
list_items_i_tpack <- list(
  "TK" = grep("^TP01_", names(data), value = T),
  "TPK" = grep("^TP02_", names(data), value = T),
  "TCK" = grep("^TP03_", names(data), value = T),
  "TPACK_ce" = grep("^TP04_(0[1-8]|12|15)$", names(data), value = T), # Sub-Subskala 1: Celik Enhanced Items
  "TPACK_genai" = grep("^TP04_(09|10|11|13|14)$", names(data), value = T), # Sub-Subskala 2: GenAI Items
  "Ethics" = grep("^TP05_0[1-6]$", names(data), value = T) #im pretest nur items 1-6 vorhanden 7,8 kamen im post dazu
)

# ein paar ethik items müssen noch rausgenommen werden, die nur im posttest erhoben wurden. hier nur check
# data_subset1 <- data %>% select(all_of(unlist(list_items_i_tpack)))
# View(data_subset1)

## Perceived Usefulnes
list_items_pu <- list(
  "Pu" = grep("^ED03_", names(data), value = T)
)
# data_subset2 <- data %>% select(all_of(unlist(list_items_pu)))
# sum(data_subset2 < 0, na.rm = TRUE)

## Reflection (bei Piloten irrelevant)
list_items_ref <- list(
  "Ref" = grep("^RE01_", names(data), value = T)
)
# data_subset3 <- data %>% select(all_of(unlist(list_items_ref)))
# sum(data_subset3 < 0, na.rm = TRUE)

## AI Literacy
# data <- data %>% mutate(AILIT = AI35_02) # rename to sumscore
list_items_ailit_wolke <- list(
  "ailit_wolke" = grep(
    #(AI05|AI06|AI14|AI15|AI17_0[1-5]|AI21|AI23|AI25|AI27|AI28|AI29)
    # "^(AI36_05|AI36_06|AI36_13|AI36_14|AI36_16|AI36_20|AI36_22|AI36_24|AI36_25|AI36_26|AI36_27)", # see excel file "Fragebogen" for decisions for each item
    "^(AI36_04|AI36_10|AI36_11|AI36_13|AI36_15|AI36_16|AI36_17|AI36_19|AI36_24|AI36_25|AI36_27)", # Sinas version
    names(data),
    value = T
  ), # items we think students learn in the WoLKE seminar
  "ailit_short" = grep(
    #(AI01|AI04|AI09|AI15|AI17_0[1-5]|AI19|AI21|AI25|AI27|AI29)
    "^(AI36_01|AI36_04|AI36_09|AI36_14|AI36_16|AI36_18|AI36_20|AI36_24|AI36_25|AI36_27)$",
    names(data),
    value = T
  ) # short version of the AILIT
)

# create new columns with the means of the lists
data <- data %>%
  mutate(
    AILIT = AI35_02, # default sumscore
    AILITWolke = rowSums(
      select(., all_of(list_items_ailit_wolke$ailit_wolke)),
      na.rm = TRUE
    ),
    AILITShort = rowSums(
      select(., all_of(list_items_ailit_wolke$ailit_short)),
      na.rm = TRUE
    ),
    TK = rowMeans(select(., all_of(list_items_i_tpack$TK)), na.rm = TRUE),
    TPK = rowMeans(select(., all_of(list_items_i_tpack$TPK)), na.rm = TRUE),
    TCK = rowMeans(select(., all_of(list_items_i_tpack$TCK)), na.rm = TRUE),
    TPACKce = rowMeans(
      select(., all_of(list_items_i_tpack$TPACK_ce)),
      na.rm = TRUE
    ),
    TPACKgenai = rowMeans(
      select(., all_of(list_items_i_tpack$TPACK_genai)),
      na.rm = TRUE
    ),
    Ethics = rowMeans(
      select(., all_of(list_items_i_tpack$Ethics)),
      na.rm = TRUE
    ),
    PU = rowMeans(select(., all_of(list_items_pu$Pu)), na.rm = TRUE),
    PU1 = ED03_01,
    PU2 = ED03_02,
    PU3 = ED03_03,
    PU4 = ED03_04,
    Ref = rowMeans(select(., all_of(list_items_ref$Ref)), na.rm = TRUE),
    TPACK_total = rowMeans(
      select(., all_of(unlist(list_items_i_tpack))),
      na.rm = TRUE
    )
  )

# check missing values in TPACK and PU scales (NA means, not answered (-1 diff. answer, -9 not answered))
data %>%
  select(all_of(c(list_items_i_tpack, list_items_pu) %>% unlist)) %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  sum

```

## Analysis

Im folgenden verwenden wir drei Kontraste: (0) Intercept = Gesamtdurchschnitt aller Gruppen, (1) Exp-Ø (MINT+Sprache)/2 vs. Control, (2) Sprache vs. MINT. Mit der Zeit-Interaktion testen wir, ob sich diese Unterschiede über die Zeit unterschiedlich entwickeln (Difference-in-Differences: Exp-Ø vs. Control sowie Sprache vs. MINT).
``` {r lmer MINT vs Sprache vs Kontrollgruppe}

data <- data %>% droplevels() # drop unused factor levels

## define contrast Control vs Exp1+Exp2 - vermutlich bessere version, da die Estimates "stimmen"
mat <- cbind(
  c(1 / 3, 1 / 3, 1 / 3), # Durchschnitt aller 3 Gruppen
  c(-1, .5, .5), # Vergleich Control vs Mittel der beiden Interventions-Gruppen
  c(0, -1, 1)
) # Vergleich Sprache vs MINT
contrasts.CvsExp <- solve(t(mat))[, -1]
colnames(contrasts.CvsExp) <- c("-C+MINT_Sprache", "-MINT+Sprache")
contrasts(data$intervention) <- contrasts.CvsExp

# or use this coding system?
# my.helmert <- matrix(c(-2 / 3, 1 / 3, 1 / 3, 0, -1 / 3, 1 / 3), ncol = 2)
# colnames(my.helmert) <- c("-C+MINT_Sprache", "-MINT+Sprache")
# contrasts(data$cond2) <- my.helmert

## define variables
random_terms_lmer <- c("(1 | code)")
groups <- c("intervention")
outcomes <- c(
  "AILIT",
  "AILITWolke",
  "AILITShort",
  "TK",
  "TPK",
  "TCK",
  "TPACKgenai",
  "TPACKce",
  "Ethics",
  "PU",
  "PU1",
  "PU2",
  "PU3",
  "PU4"
)
time <- "time"
controls <- c("age", "gender", "location")

# groups set as NA
lmers_prepost_all <-
  lapply(
    outcomes,
    fit_lmer,
    groups,
    time,
    controls,
    random_terms_lmer,
    NULL,
    data = data
  )

modelsummary_models(lmers_prepost_all)

```

```{r scale reliability with cronbach alpha}	

## look at MSD and cronbach alphas for outcomes for each intervention group
# filter out pre data and relevant lists
data_pre <- data %>% filter(time == "pre")
all_scales <- c(
  list_items_i_tpack,
  list_items_pu
)
# 1) Cronbach's alpha for each scale
alphas <- imap(
  all_scales,
  ~ psych::alpha(dplyr::select(data_pre, dplyr::all_of(.x)), warnings = FALSE)
)

# 2) Extract only std.alpha
alpha_df <- tibble(
  scale = names(alphas),
  std_alpha = map_dbl(alphas, ~ .x$total$std.alpha)
)

# 3) Descriptives (composite = row means; switch to rowSums if you want sums)
desc_df <- imap_dfr(
  all_scales,
  ~ {
    comp <- rowMeans(dplyr::select(data_pre, dplyr::all_of(.x)), na.rm = TRUE)
    tibble(
      scale = .y,
      mean = mean(comp, na.rm = TRUE),
      sd = sd(comp, na.rm = TRUE),
      min = min(comp, na.rm = TRUE),
      max = max(comp, na.rm = TRUE)
    )
  }
)

# 4) Final summary: descriptives + std.alpha
result_df <- desc_df %>%
  left_join(alpha_df, by = "scale") %>%
  arrange(scale)

result_df

# look at reflection scales' alpha in post data
data_post <- data %>% filter(time == "post")
ref_scale <- c(
  list_items_ref
)

# 1) Cronbach's alpha for each scale
alphas_post <- imap(
  ref_scale,
  ~ psych::alpha(dplyr::select(data_post, dplyr::all_of(.x)), warnings = FALSE)
)

# 2) Extract only std.alpha
alpha_post_df <- tibble(
  scale = names(alphas_post),
  std_alpha = map_dbl(alphas_post, ~ .x$total$std.alpha)
)
alpha_post_df
```

## Descriptive Plots

### Reshape data to long format

```{r reshape-long}
data_long <- data %>%
  pivot_longer(
    cols = c(all_of(outcomes)),
    names_to = "outcome",
    values_to = "score"
  )

data_long <- data_long %>%
  mutate(
    outcome = factor(outcome, levels = outcomes)
  )
```

```{r descriptive-plots}
# TPACK
p_tpack <- plot_outcome(
  data_long %>% filter(outcome %nin% c("AILIT", "PU")),
  scales_facet = "fixed"
) %+%
  ggtitle("Intelligent-TPACK")

ggsave(
  p_tpack %+% facet_wrap(~outcome, ncol = 3),
  file = "figures/plot-tpack.pdf",
  width = 8,
  height = 10,
  device = cairo_pdf
)
# AILIT and PU
p_ailit_pu <- plot_outcome(
  data_long %>% filter(outcome %in% c("AILIT", "PU"))
) %+%
  ggtitle("AI Literacy and Perceived Usefulness")
```

```{r save-env }
# save environment
save.image("data/env-analysis.Rdata")
```