---
title: "WoLKE-Pilot-2025"
author: "Benedikt Beuttler"
date: "`r Sys.Date()`"
lang: de
page-layout: full
document-css: false
link-citations: true
number-sections: true
editor: visual
format:
  html:
    standalone: true
    embed-resources: true
    section-divs: true
    html-math-method: mathjax
    wrap: none
    default-image-extension: png
    date-format: long
---

# Preamble

## Prepare the Environment

## Load Libraries

```{r load-libraries}
library(this.path)
suppressPackageStartupMessages(library(dplyr)) #e.g. for select
library(tidyr) # for reshaping data
suppressPackageStartupMessages(library(stringdist)) #levenshtein
library(stringr) # for reshaping data
```

## Load Pre- and Posttest and Reshape Data

```{r load-data}
# Set working directory to the script's location
setwd(this.path::this.dir())
### PRETEST
# load csv -> df_pre
source("./data/import_ki-lehramt_2025-07-24_08-36.r")

# extract items starting with TP (TPACK) or DE (Demographics)
data_pre <- df_pre  %>% select(QUESTNNR, starts_with("DE"), starts_with("AI"), starts_with("ED"), starts_with("TP"), starts_with("RE"), starts_with("STARTED"),  starts_with("FINISHED"))
data_pre$QUESTNNR[substr(data_pre$STARTED, 6, 7) == "07"] <- "mint-post" # fixe some wrong entries in pre list to post

# rename demographics
data_pre <- data_pre %>%
  mutate(
    code       = tolower(DE13_01),
    age        = DE14_01,
    semester   = DE15_01,
    studies    = DE16,
    subject1   = DE24,
    subject2   = DE25,
    gender     = DE10,
    location   = DE11,
    semestercount = DE15_01,
    isp        = DE09,
    sprache_control = DE08,
    sprache_last_sem = DE06, #already visited cans KI seminar
    mint_control = DE07,
    mint_exclusion = DE02
  )
data_pre <- data_pre[, c("code", setdiff(names(data_pre), "code"))] #set code as first column
# View(data_pre[(is.na(data_pre$code) | data_pre$code == ""), ])
data_pre <- data_pre[!(is.na(data_pre$code) | data_pre$code == ""), ] # remove empty/na codes (=7) -> 1 stimmt Datenschutz nicht zu

#### POSTTEST

# load csv -> df_post
source("./data/import_ki-lehramt_2025-07-24_08-35.r")

# extract items starting with
data_post <- df_post  %>% select(QUESTNNR, starts_with("DE"), starts_with("AI"), starts_with("ED"), starts_with("TP"), starts_with("RE"), starts_with("STARTED"), starts_with("FINISHED"))

# rename demographics
data_post <- data_post %>%
  mutate(
    code       = tolower(DE13_01),
    age        = DE14_01,
    semester   = DE15_01,
    studies    = DE16,
    subject1   = DE24,
    subject2   = DE25,
    gender     = DE10,
    location   = DE11,
    semestercount = DE15_01,
    isp        = DE09,
    sprache_control = DE08,
    #sprache_info = DE06, #already visited cans KI seminar
    mint_control = DE07,
    mint_exclusion = DE02
  )
data_post <- data_post[, c("code", setdiff(names(data_post), "code"))] #set code as first column
# View(data_post[(is.na(data_post$code) | data_post$code == ""), ])
data_post <- data_post[!(is.na(data_post$code) | data_post$code == ""), ] # remove empty/na codes (=8) -> 1 hat schon mint besucht, 2 stimmen nicht Datenschutz zu

# remove non basic types to make it mergeable
# Convert everything to character, then use type.convert to infer proper base types
clean_df <- function(df) {
  df[] <- lapply(df, function(col) {
    if (!is.atomic(col)) col <- as.character(col)
    col <- as.character(col)  # force avector etc. to become characters
    type.convert(col, as.is = TRUE)
  })
  return(df)
}
data_pre_clean <- clean_df(data_pre)
data_post_clean <- clean_df(data_post)

# merge pre und post dfs
#data <- dplyr::bind_rows(data_pre, data_post)
data <- bind_rows(data_pre_clean, data_post_clean)

### EXCLUSIONS
e1 <- nrow(data[data$FINISHED == "FALSE", ]) # not FINISHED
e1
data <- data[data$FINISHED == "TRUE", ] 

e2 <- nrow(data[data$code == "höel07", ])
e2
data <- data[data$code != "höel07", ] #example code

# finding typos in code
single_codes <- unique(data$code) # only one code present
# Compute pairwise Levenshtein distance
dist_matrix <- stringdist::stringdistmatrix(single_codes, single_codes, method = "lv")
# Check pairs with distance ≤ 2
for (i in seq_along(single_codes)) {
  for (j in seq_along(single_codes)) {
    if (i < j && dist_matrix[i, j] <= 2) {
      cat(sprintf("Possible match: %s <-> %s (distance = %d)\n",
                  single_codes[i], single_codes[j], dist_matrix[i, j]))
    }
  }
}
data$code[data$code == "mair10"] <- "moir10"
data$code[data$code == "drch17"] <- "frch17"
data$code[data$code == "loma26"] <- "goma26"
data$code[data$code == "rosa21"] <- "rosa20"
data$code[data$code == "sisu00"] <- "sisu01"

# create new columns with scale naming
data <- data %>%
  mutate(across(starts_with("TP01_"), identity, .names = "TPACK_TK_{str_sub(.col, -2)}")) %>%
  mutate(across(starts_with("TP02_"), identity, .names = "TPACK_TPK_{str_sub(.col, -2)}")) %>%
  mutate(across(starts_with("TP03_"), identity, .names = "TPACK_TCK_{str_sub(.col, -2)}")) %>%
  mutate(across(starts_with("TP04_"), identity, .names = "TPACK_TPACK_{str_sub(.col, -2)}")) %>%
  mutate(across(starts_with("TP05_"), identity, .names = "TPACK_Ethic_{str_sub(.col, -2)}"))

length(unique(data$code)) # only one code present
sum(table(data$code) > 1) # code paare
table(data$gender, data$QUESTNNR) # genders across pre/post in subjects
table(data$location, data$QUESTNNR) # from where across pre/post in subjects
table(data$mint_control)
table(data$code, data$QUESTNNR)

```

# Clean data
Clean up the data of each questionnaire for factor analysis. That is, we replace negative values (i.e., -1 = *did not understand the question*, -9 = no answer) with `NA`.
Disclaimer: in the "ki-lehramt" study, we do not have missing answers (-9), as all questions regarding Intelligent-TPACK were mandatory.

```{r clean data}
# todo: check values (-1, -9)
# look at ata of every questionnaire
# AI Literacy
data_ai <- data %>% select(-starts_with("ED"), -starts_with("TP"), -starts_with("RE"), -starts_with("STARTED"), -starts_with("FINISHED"),) 

# Perceived Usefulness
data_perc_use <- data %>%
  select(-starts_with("AI"), -starts_with("TP"), -starts_with("RE"), -starts_with("STARTED"),
         -starts_with("FINISHED"))
# TPACK
data_tpack <- data %>%
  select(-starts_with("AI"), -starts_with("ED"), -starts_with("RE"), -starts_with("STARTED"),
         -starts_with("FINISHED"))
# Reflection
data_ref <- data %>%
  select(-starts_with("AI"), -starts_with("ED"), -starts_with("TP"), -starts_with("STARTED"),
         -starts_with("FINISHED"))


```